---
title: "Hands-on Exercise 2A: First-order Spatial Point Pattern Analysis (1st-SPPA)"
---

## 1 Overview

**Spatial Point Pattern Analysis (SPPA)** is the evaluation of the pattern or distribution of a set of points on a surface. The points may represent:

- events such as crimes, traffic accidents, or disease onsets, or
- business services (e.g., coffee shops and fast-food outlets) or facilities such as childcare centres and eldercare centres.

**First-order Spatial Point Pattern Analysis (1st-SPPA)** focuses on understanding the intensity or density of points across a study area. It examines how the distribution of points varies over space, essentially identifying trends or patterns in point density. This type of analysis deals with the individual locations of points and their distribution, without considering interactions between them.

This exercise uses [**spatstat**](https://cran.r-project.org/web/packages/spatstat/) to perform two commonly used 1st-order SPPA methods, applied to childcare centres in Singapore, and will focus on two specific questions:

- Are the childcare centres in Singapore randomly distributed throughout the country?
- If the answer is not, then the next logical question is: where are the locations with higher concentration of childcare centres?

## 2 The Data

Two data sets are used for this exercise:

| Dataset | Format | Source | Description |
|---|---|---|---|
| Child Care Services | GeoJSON | [data.gov.sg](https://data.gov.sg/) | Point feature data providing the location and attribute information of childcare centres in Singapore. |
| Master Plan 2019 Subzone Boundary (No Sea) | GeoJSON | [data.gov.sg](https://data.gov.sg/) | Polygon feature data providing the URA 2019 Master Plan Planning Subzone boundaries. |

## 3 Installing and Loading the R packages

Four R packages will be used for this exercise:

| Package | Purpose | Use Case in Exercise |
|---|---|---|
| [**sf**](https://r-spatial.github.io/sf/) | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data in R. |
| [**spatstat**](https://spatstat.org/) | Provides tools for point pattern analysis. | Performing 1st-order spatial point pattern analysis and deriving kernel density estimation (KDE). |
| [**terra**](https://rspatial.github.io/terra/index.html) | Modern spatial data package built to replace `raster`; offers improved speed for raster and vector data. | Converting `spatstat` image output into raster format for cartographic-quality mapping. |
| [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) | Creates static and interactive thematic maps using cartographic-quality elements. | Plotting the KDE surface as a static map, and the childcare point locations as an interactive map via [leaflet](https://leafletjs.com/). |

The code chunk below installs and loads the four R packages:

```{r}
pacman::p_load(sf, spatstat, terra, tmap, tidyverse)
```

## 4 Reproducibility

As this exercise will involve Monte Carlo simulation, a seed is set here so that the result stays consistent across re-renders.

```{r}
set.seed(123)
```

