---
title: "Hands-on Exercise 2A: First-order Spatial Point Pattern Analysis (1st-SPPA)"
---

## 1 Overview

**Spatial Point Pattern Analysis (SPPA)** is the evaluation of the pattern or distribution of a set of points on a surface. The points may represent:

- events such as crimes, traffic accidents, or disease onsets, or
- business services (e.g., coffee shops and fast-food outlets) or facilities such as childcare centres and eldercare centres.

**First-order Spatial Point Pattern Analysis (1st-SPPA)** focuses on understanding the intensity or density of points across a study area. It examines how the distribution of points varies over space, essentially identifying trends or patterns in point density. This type of analysis deals with the individual locations of points and their distribution, without considering interactions between them.

This exercise uses [**spatstat**](https://cran.r-project.org/web/packages/spatstat/) to perform two commonly used 1st-order SPPA methods, applied to childcare centres in Singapore, and will focus on two specific questions:

- Are the childcare centres in Singapore randomly distributed throughout the country?
- If the answer is not, then the next logical question is: where are the locations with higher concentration of childcare centres?

## 2 The Data

Two data sets are used for this exercise:

| Dataset | Format | Source | Description |
|------------------|------------------|------------------|------------------|
| Child Care Services | GeoJSON | [data.gov.sg](https://data.gov.sg/) | Point feature data providing the location and attribute information of childcare centres in Singapore. |
| Master Plan 2019 Subzone Boundary (No Sea) | GeoJSON | [data.gov.sg](https://data.gov.sg/) | Polygon feature data providing the URA 2019 Master Plan Planning Subzone boundaries. |

## 3 Installing and Loading the R packages

Four R packages will be used for this exercise:

| Package | Purpose | Use Case in Exercise |
|------------------------|------------------------|------------------------|
| [**sf**](https://r-spatial.github.io/sf/) | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data in R. |
| [**spatstat**](https://spatstat.org/) | Provides tools for point pattern analysis. | Performing 1st-order spatial point pattern analysis and deriving kernel density estimation (KDE). |
| [**terra**](https://rspatial.github.io/terra/index.html) | Modern spatial data package built to replace `raster`; offers improved speed for raster and vector data. | Converting `spatstat` image output into raster format for cartographic-quality mapping. |
| [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) | Creates static and interactive thematic maps using cartographic-quality elements. | Plotting the KDE surface as a static map, and the childcare point locations as an interactive map via [leaflet](https://leafletjs.com/). |

The code chunk below installs and loads the four R packages:

```{r}
pacman::p_load(sf, spatstat, terra, tmap, tidyverse)
```

## 4 Reproducibility

As this exercise will involve Monte Carlo simulation, a seed is set here so that the result stays consistent across re-renders.

```{r}
set.seed(123)
```

## 5 Importing and Wrangling Geospatial Data Sets

The code chunk below imports the Master Plan 2019 Subzone Boundary (No Sea) data set, transforms it to the SVY21 projected coordinate system, and retains only the fields needed for the analysis that follows:

```{r}
mpsz_sf <- st_read("data/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414) %>%
  select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C, geometry) %>%
  relocate(geometry, .after = last_col())
```

The Southern Group subzone, and the Western Islands and North-Eastern Islands planning areas, are excluded, since these consist mainly of offshore islands with no childcare centres.

```{r}
mpsz_cl <- mpsz_sf %>%
  filter(SUBZONE_N != "SOUTHERN GROUP",
         PLN_AREA_N != "WESTERN ISLANDS",
         PLN_AREA_N != "NORTH-EASTERN ISLANDS")
```

The cleaned boundary data set is saved as an `.rds` file so it can be reloaded directly in later sections without repeating the import and filtering steps.

```{r}
write_rds(mpsz_cl, "data/mpsz_cl.rds")
```

Next, the Child Care Services data set is imported, transformed to the same coordinate reference system as `mpsz_cl`. Only the `NAME` field is retained, since address and administrative details are not required for the spatial point pattern analysis that follows.

```{r}
childcare_sf <- st_read("data/ChildCareServices.geojson") %>%
  st_transform(crs = 3414) %>%
  select(NAME, geometry)
```

::: {.callout-note}
## Why select() is needed here
KML's driver in GDAL flattens all attributes into a single `Description` HTML blob, so a KML import of this same data set would return only two fields — `Name` and `Description`. GeoJSON preserves every original attribute as its own column instead, so this data set arrives with 16 fields (e.g., address components, hyperlinks, and photo URLs). `select()` trims this down to only `NAME` and the geometry, which are the fields relevant to the spatial point pattern analysis.
:::

Before proceeding, the coordinate reference systems of both data sets are checked to confirm they match. Mismatched CRSs would silently distort every distance and area calculation in the sections that follow, so this check is a necessary safeguard.

```{r}
st_crs(mpsz_cl)
```

```{r}
st_crs(childcare_sf)
```

::: {.callout-note}
Both coordinate reference systems match exactly — both report EPSG:3414 with identical WKT parameters (SVY21 datum, Singapore Transverse Mercator projection), confirming the two data sets share a consistent spatial reference.
:::

As a final check, both data sets are confirmed to remain in `sf` format after the import and wrangling steps above:

```{r}
class(mpsz_cl)
class(childcare_sf)
```

### 5.1 Mapping the Geospatial Data Sets

With both `mpsz_cl` and `childcare_sf` confirmed to share the same coordinate reference system, a static map is prepared below to visualise their spatial patterns together.

```{r}
tm_shape(mpsz_cl) +
  tm_polygons() +
tm_shape(childcare_sf) +
  tm_dots()
```

::: {.callout-tip}
## Reading the code

`tm_shape()` declares which dataset to draw, and the layer function that follows it determines how: `tm_polygons()` renders `mpsz_cl` as filled subzone boundaries, while `tm_dots()` renders `childcare_sf` as point markers. The `+` operator layers them onto the same map, boundaries first so the childcare points render on top.
:::

::: {.callout-important}
Both geospatial layers fall within the same map extent, confirming their coordinate values refer to the same spatial context. This is an important check before any geospatial analysis proceeds.
:::

Alternatively, an interactive point symbol map can also be prepared using the code chunk below:

```{r}
#| message: false
tmap_mode('view')
tm_shape(childcare_sf) +
  tm_dots()
```

::: {.callout-note}
In interactive mode, **tmap** uses the [leaflet for R](https://leafletjs.com/) API. This allows the map to be freely navigated and zoomed, and the attributes of each point queried by clicking on it. The background map layer can also be changed — three options are provided (ESRI.WorldGrayCanvas, OpenStreetMap, and ESRI.WorldTopoMap), with **ESRI.WorldGrayCanvas** as the default.
:::

```{r}
#| message: false
tmap_mode('plot')
```

::: {.callout-warning}
## Switch back to plot mode

Remember to switch back to plot mode after viewing the interactive map. Each interactive map consumes a connection, and displaying too many interactive maps (more than 10) in a single document can cause issues when published on Netlify.
:::

## 6 Geospatial Data Wrangling

**spatstat** relies on its own specific data structures: **ppp** (planar point pattern) for point data, and **owin** for observation windows. This section converts `mpsz_cl` and `childcare_sf` from sf objects into ppp and owin objects.

### 6.1 Converting sf Data Frames to ppp Class

**spatstat** requires point event data in `*ppp*` object form. `as.ppp()` converts `childcare_sf` into this format:

```{r}
childcare_ppp <- as.ppp(childcare_sf)
```

Next, `class()` verifies the object class of `childcare_sf`:

```{r}
class(childcare_ppp)
```

We can also take a quick look at the summary statistics of the newly converted ppp object by using `summary()`.

```{r}
summary(childcare_ppp)
```

::: {.callout-tip}
## Mark variables: NAME only
`childcare_sf` was trimmed to `NAME` and geometry back in section 5, since `DESCRIPTION` held no distinguishing information (the same constant string across all records) and the remaining address/hyperlink fields were irrelevant to spatial analysis. As a result, `NAME` is the only mark variable reported.
:::

::: {.callout-important}
The output above flags that the pattern contains duplicated points. Duplicated points are a common issue in spatial point pattern analysis, and worth taking seriously rather than glossing over. Many of the statistical methods used to analyse point patterns assume that points are distinct and non-coincident; when this assumption does not hold, the results can be misleading. Hence, this issue will be investigated below.
:::

### 6.2 Checking for Duplicated Points

The `summary()` output above flags that `childcare_ppp` contains duplicated points. In **spatstat**, this refers to points sharing identical coordinates, not identical names. `duplicated()` is the function used to check for this, and it takes an argument called `rule`, which controls exactly what counts as a duplicate:

- `rule = spatstat` (the default): a point counts as duplicated only if both its coordinates and its marks match an earlier point exactly.
- `rule = deldir`: coordinates are compared using the same method as the **deldir** package, for consistency with functions like Delaunay triangulation that rely on it.
- `rule = unmark`: only coordinates are compared, ignoring marks entirely.

```{r}
rules <- c("spatstat", "deldir", "unmark")

duplicate_counts <- list()
for (rule in rules) {
  duplicate_counts[[rule]] <- sum(duplicated(childcare_ppp, rule = rule))
}

print(duplicate_counts)
```

::: {.callout-note}
Under the strict `spatstat` rule, only **1 duplicate** is found: two entries both named "Summer Montessori" registered at the exact same coordinates, which likely reflects the same centre appearing twice in the source data rather than two distinct centres. The `deldir` rule returns the same result, since it also compares coordinates using an equivalent matching logic.

The `unmark` rule tells a very different story: **187** points share their coordinates with another point when names are ignored. A closer inspection of these locations shows that one address hosts "Childcare @ The Enrichment", "MindChamps Preschool @ Jurong West", and "Safari House Preschool JW" together, while another pairs "Key Elementary Eduhouse" with a branch of "PCF Sparkletots Preschool". These are not data-entry errors: they reflect real childcare centres, under different names and likely different operators, sharing the same building or block.

This distinction matters for how the duplication should be handled. A single true duplicate (the "Summer Montessori" case) could reasonably be removed as a data error. However, treating all 187 coordinate-sharing points as errors would be wrong, since it would delete genuinely distinct childcare centres simply because they operate from a shared address.
:::

::: {.callout-tip}
## Why the results differ sharply by rule
Under the strict `spatstat` rule, almost no points qualify as duplicates, since two centres would need identical coordinates **and** identical names. Under `unmark`, which checks coordinates alone, the count rises substantially. This gap reveals something specific about the data: many pairs of childcare centres share a building or address, but operate under different names — most likely different operators registered at the same location, rather than a single centre entered twice.
:::

The `unmark` rule above shows that **187** points share a location with another point, but not how many points cluster at each shared address.

To exactly see which centres are involved, the coincident points can be extracted and inspected directly:

```{r}
coincident_points <- duplicated(childcare_ppp, rule = "unmark")
coincident_coordinates <- childcare_ppp[coincident_points]
coincident_coordinates
```

### 6.3 Resolving the Duplicates

Coincident points can be handled in a few different ways:

| Approach | What it does |
|---|---|
| **Delete the duplicates** | Simplest option, but discards potentially valid, distinct centres |
| **Attach duplicate count as a mark** | Uses analytical methods that explicitly account for it |
| **Apply jittering** | Small random perturbation, so no two points occupy the exact same location |

As identified earlier, the "Summer Montessori" pair shares both coordinates and a name, and most plausibly reflects the same centre entered twice in the source data, rather than two different centres. The remaining coordinate-sharing groups are a different case: distinct centres, under different names, that genuinely share a building or address. Treating both cases the same way would either wrongly discard real data, or wrongly preserve a data-entry error, so the two are handled separately.

**Step 1: Removing the true duplicate**

The true duplicate is removed first, using `unique()`, which by default removes a point only if both its coordinates and its marks exactly match a point that appeared earlier:

```{r}
childcare_ppp_unique <- unique(childcare_ppp)
childcare_ppp_unique
```

This drops the point count from 1,925 to 1,924. 

::: {.callout-note}
Only the second "Summer Montessori" entry is removed; the first "Summer Montessori" entry, along with the other two distinct centres sharing that same address ("Pat's Schoolhouse" and "Star Learners @ Yung Ho") are all kept.
:::

**Step 2: Resolving the remaining coincident points**

With the removal of the true duplicate, some locations still hold more than one point. `rjitter()` resolves this by applying a small, independent random displacement to every point in the pattern, so that no two points occupy the exact same location:

```{r}
childcare_ppp_jit <- rjitter(childcare_ppp_unique,
                              retry = TRUE,
                              nsim = 1,
                              drop = TRUE)
```

To confirm that no coincident points remain after jittering:

```{r}
any(duplicated(childcare_ppp_jit))
```

From this point onward, `childcare_ppp_jit` is used in place of `childcare_ppp` for the remainder of this exercise.

### 6.4 Creating owin object

When analysing spatial point patterns, it is a good practice to confine the analysis to a defined geographical area, such as Singapore's boundary. In **spatstat**, an object called **owin** is specially designed to represent this polygonal region.

The code chunk below, `as.owin()` is used to covert `mpsz_cl` into owin object.

```{r}
sg_owin <- as.owin(mpsz_cl)
class(sg_owin)
```

The result above confirms that `sg_owin` is indeed an owin object, which can be displayed using `plot()` function:

```{r}
plot(sg_owin)
```

### 6.5 Combining the Point Events and owin Objects

In the final step of data preparation, `childcare_ppp_jit` is clipped to `sg_owin`, keeping only childcare centres located within Singapore's boundary:

```{r}
childcareSG_ppp <- childcare_ppp_jit[sg_owin]
```

The output object combined both the point and polygon feature in one ppp object class as shown below.

```{r}
childcareSG_ppp
```

::: {.callout-note}
**Observation:** From the result, it was noted that the point count drops from 1,924 to 1,923. A possible reason could be because `rjitter()` displaces each point within `childcare_ppp_jit`'s own rectangular window, which extends slightly beyond Singapore's actual coastline. One point, likely sitting close to the original boundary, was jittered into that gap between the rectangle and the true polygon, landing just outside `sg_owin`.
:::

## 7 Clark-Evans Test for Nearest Neighbour Analysis

Nearest Neighbor Analysis (NNA) is a spatial statistics method that calculates the average distance between each point and its closest neighbor to determine if a pattern of points is clustered, dispersed, or randomly distributed.

Clark-Evans test is a specific statistical method used within NNA to quantify whether a point pattern is clustered, random, or uniformly spaced, using the Clark-Evans aggregation index (R) to describe this pattern. NNA provides a numerical value that describes the degree of clustering or regularity, and the Clark-Evans test calculates a specific index (R) for this purpose.

In this section, the Clark-Evans test of aggregation for a spatial point pattern is performed using `clarkevans.test()` of the [**spatstat.explore**](https://cran.r-project.org/web/packages/spatstat.explore/refman/spatstat.explore.html) package.

The test hypothesis are:

- H0: The distribution of childcare centres is randomly distributed.
- H1: The distribution of childcare centres is not randomly distributed.

A 95% confidence interval is used.

`clarkevans.test()` of **spatstat.explore** supports two variants of the Clark-Evans test: 1. Without CSR and 2. With CSR. 

### 7.1 Performing the Clark-Evans Test Without CSR

The code chunk below uses the method without CSR:

```{r}
clarkevans.test(childcareSG_ppp,
                 correction = "none",
                 clipregion = "sg_owin",
                 alternative = c("clustered"))
```

::: {.callout-note}
## Statistical Conclusion
With R = 0.55518 (R < 1) and a p-value lower than the 0.05 level of significance, the null hypothesis is rejected. Childcare centres in Singapore are significantly clustered rather than randomly distributed.
:::

::: {.callout-tip}
## Takeaway from statistical analysis
Childcare centres in Singapore are not spread out evenly across the island. Instead, they tend to cluster together in certain areas, likely reflecting where families and young children actually live. This pattern is strong and statistically conclusive, not a result of chance variation in the data.
:::

### 7.2 Performing the Clark-Evans Test with CSR

The code chunk below uses `method = "MonteCarlo"`. In this case, the p-value is computed by comparing the observed value of R to the results from `nsim` (i.e. 39, 99, 999) simulated realisations of Complete Spatial Randomness, conditional on the observed number of points:

```{r}
clarkevans.test(childcareSG_ppp,
                 correction = "none",
                 clipregion = "sg_owin",
                 alternative = c("clustered"),
                 method = "MonteCarlo",
                 nsim = 99)
```

::: {.callout-note}
## Statistical conclusion
With R = 0.55518 (R < 1) and a p-value of 0.01, below the 0.05 level of significance, the null hypothesis is rejected. This result is consistent with the earlier test without CSR, confirming that childcare centres in Singapore are significantly clustered rather than randomly distributed.
:::

::: {.callout-tip}
## Takeaway from statistical analysis
Two independent methods (1. Without CSR and 2. With CSR) arrive at the same conclusion: childcare centres are not spread evenly across Singapore, but cluster together in certain areas. This consistency across methods adds confidence that the pattern is real, rather than an artefact of how the test was calculated.
:::

## 8 Kernel Density Estimation Method

**Kernel Density Estimation (KDE)** is a valuable tool for visualising and analysing first-order spatial point patterns. It is widely considered a method within **Exploratory Spatial Data Analysis (ESDA)**, since it transforms discrete point data (such as childcare centre locations, crime incidents, or disease cases) into continuous density surfaces that reveal clusters and variations in event occurrence, without making prior assumptions about the underlying data distribution. It helps in understanding data distribution, identifying hotspots, and exploring relationships between spatial variables before more rigorous analysis is performed.

This section covers computing the kernel density estimation (KDE) of childcare centres in Singapore.

### 8.1 Working with Automatic Bandwidth Selection Method

The code chunk below computes a kernel density using the following configurations of `density()` of **spatstat**:

- `bw.diggle()` as the automatic bandwidth selection method. Other recommended methods are `bw.CvL()`, `bw.scott()`, or `bw.ppl()`.
- The smoothing kernel used is `gaussian`, which is the default. Other smoothing methods are `epanechnikov`, `quartic`, or `disc`.
- The intensity estimate is corrected for edge effect bias using the method described by Jones (1993) and Diggle (2010, equation 18.9). The default is **FALSE**.

```{r}
kde_SG_diggle <- density(childcareSG_ppp,
                          sigma = bw.diggle,
                          edge = TRUE,
                          kernel = "gaussian")
```

::: {.callout-note}
The output of `density()` of **spatstat** is an **im** class, representing a two-dimensional pixel image. It is used to store and manipulate raster data, where the spatial domain is divided into a grid of rectangular pixels. Each pixel has an associated value, which can be numerical or a factor.
:::

The `plot()` function of Base R is then used to display the kernel density derived.

```{r}
plot(kde_SG_diggle)
```

Below is a summary report of the derived kernel density surface:

```{r}
summary(kde_SG_diggle)
```

::: {.callout-important}
The `range` above shows a maximum density of roughly 0.0000321 points per square metre. This is not because the pattern is sparse; it is a consequence of measuring density in SVY21's base unit, metres. Singapore's land area is roughly 700 million square metres, hence any density expressed per square metre produces very small decimal values regardless of how concentrated the actual points are. When expressed as points per square kilometre (multiplying by 1,000,000), the same maximum value reads as approximately 32.1, a more intuitive figure.
:::

Before moving to the next section, the bandwidth used to compute the KDE layer can be retrieved using the code below.

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

### 8.2 Rescaling KDE Values

`rescale.ppp()` converts the unit of measurement from metres to kilometres:

```{r}
childcareSG_ppp_km <- rescale.ppp(childcareSG_ppp, 1000, "km")
```

The rescaled data set is then passed into `density()` to recompute the kernel density surface, and `plot()` displays the result.

```{r}
#| fig-width: 8
#| fig-height: 6
kde_childcareSG_km <- density(childcareSG_ppp_km,
                               sigma = bw.diggle,
                               edge = TRUE,
                               kernel = "gaussian")

plot(kde_childcareSG_km)
```

The output image looks identical to the earlier version; only the data values change, as reflected in the legend.

::: {.callout-tip}
## Resolving the earlier scale issue
Rescaling to kilometres directly addresses the readability problem flagged earlier: with distances measured in kilometres rather than metres, the resulting density values land in an easily interpreted range (the legend here runs from 0 to roughly 30), instead of the near-unreadable values per square metre.
:::

### 8.3 Working with Different Automatic Bandwidth Methods

Besides `bw.diggle()`, three other **spatstat** functions can be used to determine the bandwidth: `bw.CvL()`, `bw.scott()`, and `bw.ppl()`.

The bandwidth values returned by each method are compared below:

```{r}
bw.diggle(childcareSG_ppp_km)
```

```{r}
bw.CvL(childcareSG_ppp_km)
```

```{r}
bw.scott(childcareSG_ppp_km)
```

```{r}
bw.ppl(childcareSG_ppp_km)
```

Baddeley et al. (2016) suggested the use of the `bw.ppl()` algorithm because past experience showed that it tends to produce more appropriate values when the pattern consists predominantly of tight clusters. They also noted that if the purpose of the study is to detect a single tight cluster in the midst of random noise, the `bw.diggle()` method may work better.

The code chunk below compares the output of `bw.diggle()` and `bw.ppl()` methods:

```{r}
#| fig-width: 10
#| fig-height: 8
kde_childcareSG_ppl <- density(childcareSG_ppp_km,
                                sigma = bw.ppl,
                                edge = TRUE,
                                kernel = "gaussian")

par(mfrow = c(1, 2))
plot(kde_childcareSG_km, main = "bw.diggle")
plot(kde_childcareSG_ppl, main = "bw.ppl")
```

### 8.4 Working with Different Kernel Methods

By default, the kernel method used in `density.ppp()` is `gaussian`. There are also three other options available: `epanechnikov`, `quartic`, and `disc`.

The code chunk below computes three more kernel density estimations using these three kernel functions:

```{r}
#| fig-width: 10
#| fig-height: 10
par(mfrow = c(2, 2))
plot(density(childcareSG_ppp_km,
             sigma = 0.2889242,
             edge = TRUE,
             kernel = "gaussian"),
     main = "Gaussian")
plot(density(childcareSG_ppp_km,
             sigma = 0.2889242,
             edge = TRUE,
             kernel = "epanechnikov"),
     main = "Epanechnikov")
plot(density(childcareSG_ppp_km,
             sigma = 0.2889242,
             edge = TRUE,
             kernel = "quartic"),
     main = "Quartic")
plot(density(childcareSG_ppp_km,
             sigma = 0.2889242,
             edge = TRUE,
             kernel = "disc"),
     main = "Disc")
```

## 9 Fixed and Adaptive KDE

### 9.1 Computing KDE Using a Fixed Bandwidth

A KDE layer is computed using a fixed bandwidth of 600 metres. Since the unit of measurement for `childcareSG_ppp_km` is in kilometres (600m = 0.6km), this is expressed as **sigma = 0.6**.

```{r}
kde_childcareSG_fb <- density(childcareSG_ppp_km,
                               sigma = 0.6,
                               edge = TRUE,
                               kernel = "gaussian")

plot(kde_childcareSG_fb)
```

### 9.2 Computing KDE Using Adaptive Bandwidth

The fixed bandwidth method is very sensitive to highly skewed distributions of spatial point patterns across geographical units, such as urban versus rural areas. One way to address this is to use **adaptive bandwidth**.

This section covers deriving adaptive kernel density estimation using `adaptive.density()` of **spatstat**.

```{r}
kde_childcareSG_ab <- adaptive.density(childcareSG_ppp_km,
                                        method = "kernel")

plot(kde_childcareSG_ab)
```

::: {.callout-tip}
## Why adaptive bandwidth handles this differently
A fixed bandwidth applies the same smoothing radius everywhere on the map, regardless of local point density. In a densely packed area, this can over-smooth and blur together clusters that are genuinely distinct; in a sparse area, the same fixed radius may be too narrow to reveal any pattern at all. 

Adaptive bandwidth adjusts the radius locally: shorter in dense regions, longer in sparse ones, allowing it to represent both mature, densely built-up estates and newer, sparser ones more accurately within the same map.
:::

The fixed and adaptive kernel density estimation outputs are compared using the code chunk below:

```{r}
#| fig-width: 10
#| fig-height: 8
par(mfrow = c(1, 2))
plot(kde_childcareSG_fb, main = "Fixed bandwidth")
plot(kde_childcareSG_ab, main = "Adaptive bandwidth")
```

## 10 Plotting Cartographic Quality KDE Maps

### 10.1 Converting Gridded Output into Raster

The `im` kernel density objects are converted into **SpatRaster** objects using `rast()` of the **terra** package:

```{r}
kde_childcareSG_bw_terra <- rast(kde_childcareSG_km)
```

`class()` is used to verify that `kde_childcareSG_bw_terra` belongs to the **SpatRaster** class:

```{r}
class(kde_childcareSG_bw_terra)
```

Next, the properties of `kde_childcareSG_bw_terra` are shown below:

```{r}
kde_childcareSG_bw_terra
```

::: {.callout-note}
Notice that the `coord. ref.` field above is **empty**. The reason is because `rast()` carries over the numeric grid and its extent from the `im` object, but not the coordinate reference system information. Hence, this is addressed in the next step.
:::

### 10.2 Assigning Projection Systems

In the code chunk below, `crs()` of **terra** is used to assign CRS information to the `kde_childcareSG_bw_terra` layer.

```{r}
crs(kde_childcareSG_bw_terra) <- "EPSG:3414"
```

The properties of `kde_childcareSG_bw_terra` are checked again below, to confirm the CRS assignment:

```{r}
kde_childcareSG_bw_terra
```

The `coord. ref.` field now shows "SVY21 / Singapore TM (EPSG:3414)", resolving the empty field flagged in the previous step.

### 10.3 Plotting KDE Map with tmap

Finally, the raster is displayed as a cartographic quality map using **tmap**:

```{r}
#| message: false
tm_shape(kde_childcareSG_bw_terra) +
  tm_raster(col.scale = tm_scale_continuous(values = "viridis"),
            col.legend = tm_legend(
              title = "Density values",
              title.size = 0.7,
              text.size = 0.7,
              bg.color = "white",
              bg.alpha = 0.7,
              position = tm_pos_in("right", "bottom"),
              frame = TRUE)) +
  tm_graticules(labels.size = 0.7) +
  tm_compass() +
  tm_layout(scale = 1.0)
```

The raster values are encoded explicitly onto the raster pixels using the values in the `lyr.1` field.



