---
title: "Hands-on Exercise 4: Spatial Weights and Applications"
date: "September 14, 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1 Overview

This exercise covers computing spatial weights in R and using them to derive spatially lagged variables.

A spatial weights matrix is the rule that decides which areal units count as neighbours of which, and how much weight each neighbour carries. Every measure of spatial autocorrelation is calculated from that rule, so changing it can change the result: a pattern that appears clustered under one definition of neighbourhood may not under another. The choice is part of the analysis rather than a step that precedes it.

The following will be covered during the exercise:

- importing geospatial data with **sf** package,
- importing an aspatial csv file with **readr** package,
- performing a relational join with **dplyr** package,
- computing contiguity-based, distance-based, and inverse-distance spatial weights with **spdep** package, and
- deriving spatially lagged variables with **spdep** package.

## 2 The Study Area and Data

Two data sets are used:

| Data set | Format | Description |
|------------------------|------------------------|------------------------|
| Hunan | ESRI shapefile | County boundary layer for Hunan province |
| Hunan_2012.csv | csv | Selected local development indicators for Hunan's counties in 2012 |

### 2.1 Getting Started

The code chunk below loads the packages required for this exercise:

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse, knitr)
```

## 3 Getting the Data into R

### 3.1 Importing the shapefile

The county boundary layer is imported with *st_read()* of **sf**:

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

```{r}
class(hunan)
```

### 3.2 Importing the csv file

The 2012 development indicators are imported with *read_csv()* of **readr**:

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

```{r}
class(hunan2012)
```

The file contains 88 rows and 29 columns.

### 3.3 Performing relational join

Before joining, the columns available in each source are inspected:

```{r}
colnames(hunan)
```

```{r}
colnames(hunan2012)
```

The join is performed with *left_join()* of **dplyr**, matched on `County`:

```{r}
hunan_joined <- left_join(hunan, hunan2012, by = "County")
colnames(hunan_joined)
```

```{r}
hunan <- hunan_joined %>%
  select(NAME_2, ID_3, NAME_3, ENGTYPE_3, County, GDPPC)
```

::: callout-note
Named columns are used here instead of position. `select(1:4, 7, 15)` produces the same result for these particular files, but it depends on the exact column order of both `hunan` and `hunan2012` remaining unchanged. If either source were reordered or extended, the same code would silently select a different column with no error, so naming the columns removes that dependency.
:::

## 4 Visualising Regional Development Indicator

A basemap with county labels and a choropleth of GDPPC are prepared with **tmap**.

```{r}
#| fig-width: 12
#| fig-height: 9
basemap <- tm_shape(hunan) +
  tm_polygons(fill = "grey90") +
  tm_text("NAME_3", size=0.5)

gdppc <- qtm(hunan, "GDPPC") +
  tm_layout(
    legend.position = c("left", "bottom"))
tmap_arrange(basemap, gdppc, asp=1, ncol=2)
```

::: callout-note
**Interpretation:**

Counties are shaded from light blue to dark blue according to GDP per capita, with darker shades indicating higher values. The resulting pattern shows where economic activity is concentrated across Hunan's counties.

Most of the province sits in the lowest bracket. The dominant shade is the lightest blue (0–19,999), indicating that the bulk of Hunan's counties cluster at the low end of the GDPPC range. Wealth is concentrated rather than evenly spread.

The darker cluster (60,000–79,999 and 80,000-100,000) sits in the northeast, and it is a tight group of adjacent counties rather than a single outlier. Changsha has the highest GDPPC in the Hunan province.

- **One hotspot:** Changsha, Wangcheng, Liuyang, Ningxiang
- **Two separate outliers:** Zixing and Lengshuijiang
:::

## 5 Computing Contiguity Spatial Weights

[poly2nb()](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package can be used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries. The `queen` argument controls what counts as sharing a boundary: under Queen contiguity a single shared vertex is sufficient, whereas Rook contiguity requires a shared edge. The argument defaults to **TRUE**, so calling the function without specifying queen returns Queen contiguity neighbours.

### 5.1 Computing QUEEN contiguity

```{r}
wm_q <- poly2nb(hunan, queen = TRUE)
summary(wm_q)
```

```{r}
hunan$County[c(30, 65, 85)]
```

::: callout-note
**Interpretation:**

All 88 counties have at least one neighbour, so no county is spatially isolated under the Queen criterion. The number of non-zero links is 448, with an average of 5.09 links per county. The link distribution is concentrated in the middle of its range, with most counties having between three and seven neighbours.

Taoyuan is the most connected county with 11 neighbours, while Xinhuang and Linxiang have the least connected counties, each with a single neighbour.
:::

The neighbours of Taoyuan can be examined using the code below:

```{r}
wm_q[[85]]
```

The corresponding county names:

```{r}
hunan$NAME_3[wm_q[[85]]]
```

::: callout-note
The neighbour identifiers are row positions in `hunan`, not county codes, so they need to be resolved against the data frame before being interpreted.
:::

The GDPPC of the above 11 neighbouring counties:

```{r}
nb85 <- hunan$GDPPC[wm_q[[85]]]
nb85
```

The structure of the full neighbours list can be inspected with `str()`.

```{r}
str(wm_q)
```

### 5.2 Computing ROOK contiguity

```{r}
wm_r <- poly2nb(hunan, queen = FALSE)
summary(wm_r)
```

```{r}
hunan$County[c(30, 65, 85)]
```

```{r}
hunan$NAME_3[wm_r[[85]]]
```

**Nan** is excluded from Taoyuan's neighbours under Rook contiguity.

::: callout-note
**Interpretation:**

All 88 counties have at least one neighbour under Rook contiguity, so no county becomes isolated when point-only contacts are excluded. The number of non-zero links is 440, with an average of 5.0 links per county. The link distribution is concentrated in the middle of its range, with most counties having between three and seven neighbours.

Taoyuan is the most connected county with 10 neighbours, while Xinhuang and Linxiang have the least connected counties, each with a single neighbour.
:::

### 5.3 Comparing Queen and Rook contiguity

Rook contiguity produces fewer links than Queen, with an average of 5.0 links per county compared to 5.09 under Queen. This is expected, since every Rook neighbour is also a Queen neighbour, a shared edge always implies at least a shared vertex, but the reverse does not hold.

The county-level neighbour lists make this concrete: Taoyuan has 11 neighbours under Queen and 10 under Rook, losing exactly one, Nan. Checking the shared geometry between the two counties confirms the contact is a single point rather than an edge, so it satisfies Queen contiguity but not Rook's stricter requirement, exactly the case the two criteria are meant to distinguish. Xinhuang and Linxiang remain least connected under both criteria, showing that their isolation reflects their position on the map rather than an artefact of which contiguity rule is applied.

### 5.4 Visualising contiguity weights

A connectivity graph draws a line between points, not between whole polygon shapes, so each polygon first needs a single point to represent it. A centroid, the geometric centre of a shape, serves this purpose. Simply calling `st_centroid()` on the geometry column would return sf point objects, but the connectivity graph's plotting function needs plain coordinate values in a two-column table instead, so an extra step is needed to get from one format to the other.

To do this, a mapping function is used: `map_dbl()` from **purrr**. A mapping function takes a function and applies it separately to every item in a list, then collects all the results together into one vector. Here, the list being worked through is hunan's geometry column, which holds one polygon shape per county, and the function applied to each shape is `st_centroid()`.

Longitude is extracted first:

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])
```

::: callout-note
`.x` stands for whichever polygon is currently being processed as *map_dbl()* works through the geometry column one element at a time. *st_centroid(.x)* computes that polygon's centroid, and `[[1]]` retrieves the first coordinate value, longitude, from the result.
:::

Latitude is obtained the same way, differing only in which coordinate is retrieved:

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

::: callout-note
`[[2]]` retrieves the second coordinate value, latitude, from each centroid.
:::

The two vectors are then combined into a single object with `cbind()`:

```{r}
coords <- cbind(longitude, latitude)
```

The first few rows are checked to confirm the result is formatted correctly:

```{r}
head(coords)
```

### 5.5 Plotting Queen and Rook Contiguity Maps

Queen and Rook contiguity are plotted side by side:

```{r}
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "Queen Contiguity")
plot(wm_q, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
plot(hunan$geometry, border = "lightgrey", main = "Rook Contiguity")
plot(wm_r, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

```{r}
#| include: false
par(mfrow = c(1, 1))
```

The two graphs are visually near-identical, consistent with the small difference in link counts reported above.

## 6 Computing Distance-based Neighbours

[`dnearneigh()`](https://search.r-project.org/CRAN/refmans/spdep/html/dnearneigh.html) of **spdep** identifies neighbouring region points based on their Euclidean distance, controlled by two separate arguments: a lower bound (d1) and an upper bound (d2).

If the coordinates are not projected (i.e., in latitude and longitude), whether provided as a spatial coordinates object or as a plain two-column matrix, with longlat = TRUE, the function will calculate great circle distances in kilometres, assuming the WGS84 reference ellipsoid.

### 6.1 Determining the cut-off distance

Firstly, the upper limit for distance band needs to be determined. This can be done through the following steps:

- Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [knearneigh()](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.
- Convert the knn object returned by `knearneigh()` into a neighbours list of class **nb** with a list of integer vectors containing neighbour region number ids by using [knn2nb()](https://r-spatial.github.io/spdep/reference/knn2nb.html).
- Return the length of neighbour relationship edges by using [nbdists()](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. If the coordinates are projected, the function returns distances in the units of the coordinates; otherwise, in km.
- Remove the list structure of the returned object by using [unlist()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

::: {.callout-note}
The above output shows that the maximum first-neighbour distance is 61.79km. Using this value as the upper threshold gives certainty that all units will have at least one neighbour.
:::

### 6.2 Computing fixed distance weight matrix

The maximum of the first-nearest-neighbour distances calculated above is the distance from the single most isolated county to its own nearest neighbour.

```{r}
cutoff <- ceiling(max(k1dists))
cutoff
```

The derived cut-off is applied as the upper bound of the distance band, with the lower bound at 0:

```{r}
wm_d <- dnearneigh(coords, 0, cutoff, longlat = TRUE)
wm_d
```

::: {.callout-note}
**Interpretation:**

The output shows that the average number of links is 3.681818 (the number of nonzero links (324) divided by the number of counties (88)). This means that, on average, each county has roughly 3.68 other counties within 62 km of its centroid.

This is noticeably lower than the 5.09 average links per county found under Queen contiguity. The comparison is meaningful because both are describing neighbourhood size for the same 88 counties, just defined in two different ways. Queen contiguity treats two counties as neighbours if their boundaries touch at all, no matter how long or short the shared edge is. The fixed-distance band works differently as it only counts a neighbour if the two centroids are close enough, regardless of whether the counties share a border. Hence, this is why some Queen neighbours do not qualify. A large county can border several smaller ones along its edge, but its centroid might sit far from theirs, too far to fall within the distance band, even though the two counties are clearly touching on the map.
:::

The structure of the neighbours list can be inspected using `str()`:

```{r}
str(wm_d)
```

The distribution of link counts by county is cross-tabulated with [table()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table) and [card()](https://r-spatial.github.io/spdep/reference/card.html), where `card()` returns the number of neighbours for each element in the list. Each row of the table corresponds to one county, and each column corresponds to a neighbour count observed across the province.

```{r}
table(hunan$County, card(wm_d))
```

The neighbours list is checked for disconnected subgraphs with `n.comp.nb()`, which groups counties into connected components: a set of counties is one component if every county in it can be reached from every other through a chain of neighbour links, else a separate component.

```{r}
n_comp <- n.comp.nb(wm_d)
n_comp$nc
```

```{r}
table(n_comp$comp.id)
```

::: {.callout-note}
**Interpretation:**

The output confirms a single component of all 88 counties, so the neighbours list is fully connected. Every county can be reached from every other through a chain of neighbour relationships.
:::

#### 6.2.1 Plotting fixed distance weight matrix

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(wm_d, coords, add = TRUE)
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
```

Red lines mark the 1st nearest neighbour links and black lines mark all links within the cut-off distance of 62km. The contrast shows how much denser the neighbourhood structure becomes once the full distance band is admitted.

```{r}
par(mfrow = c(1, 2))
plot(hunan$geometry, border = "lightgrey", main = "1st nearest neighbours")
plot(k1, coords, add = TRUE, col = "red", length = 0.08)
plot(hunan$geometry, border = "lightgrey", main = "Distance link")
plot(wm_d, coords, add = TRUE, pch = 19, cex = 0.6)
```

```{r}
#| include: false
par(mfrow = c(1, 1))
```

::: {.callout-note}
**Interpretation:**

The two networks are not independent. This is because the 62 km cut-off was derived as the ceiling of the largest 1st nearest neighbour distance, so every 1st nearest neighbour link is guaranteed to fall within the fixed distance band. The red links is therefore a subset of the black links.
:::

### 6.3 Computing adaptive distance weight matrix

One of the characteristics of fixed distance weight matrix is that more densely settled areas (usually the urban areas) tend to have more neighbours while the less densely settled areas (usually the rural counties) tend to have lesser neighbours. This matters because a county's lagged value is the average of its neighbours' values. Averaging over more neighbours smooths that average, diluting the influence of any single neighbour's unusually high or low value.

It is possible to control the numbers of neighbours directly using **k-nearest neighbours**, either accepting asymmetric neighbours or imposing symmetry as shown in the code below.

```{r}
knn6 <- knn2nb(knearneigh(coords, k = 6))
knn6
```

::: {.callout-note}
**Interpretation:**

The output reports a non-symmetric neighbours list. This highlights that if county A is among county B's six nearest neighbours, it does not necessarily mean that B is among A's six nearest. This property is intrinsic to the adaptive scheme and does not arise with contiguity or fixed distance bands (both of which are symmetric by construction).
:::

The structure of the neighbours list can be inspected using `str()`:

```{r}
str(knn6)
```

**Each county has exactly six neighbours.**

#### 6.3.1 Plotting adaptive distance weight matrix

```{r}
plot(hunan$geometry, border = "lightgrey")
plot(knn6, coords, pch = 19, cex = 0.6, add = TRUE, col = "red")
```

As every county is linked to exactly six others, the network is visibly denser and more evenly distributed compared to the fixed distance network.

## 7 Weights Based on Inverse Distance Weighting (IDW)

Every method used so far treats neighbours as either fully counted or not counted at all. Under Queen contiguity, a county either touches another county or it does not. There is no partial credit for touching along a longer border versus a shorter one. Under a fixed distance band, a county within 62 km counts fully as a neighbour and a county 63 km away counts not at all.

Inverse distance weighting takes a different approach: Instead of a strict in-or-out rule, it lets a neighbour's influence shrink gradually as the distance to it grows. A close neighbour gets a large weight, a far neighbour gets a small weight, and there is no sharp cutoff where influence suddenly drops to zero.

This reflects an idea often called the first law of geography: nearby things tend to be more closely related than distant things. Inverse distance weighting builds that assumption directly into the numbers, rather than relying on a fixed boundary to decide who counts as a neighbour.

The distances between Queen contiguity neighbours are computed with [nbdists()](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**.

```{r}
nb_dist <- nbdists(wm_q, coords, longlat = TRUE)
ids <- lapply(nb_dist, function(x) 1 / x)
ids
```

## 8 Row-standardised Weights Matrix

Each neighbouring county now needs to be assigned a weight. Under **style = "W"**, every neighbour is treated as equally important: each one receives an equal share of the total weight **(1 divided by number of neighbours)**. A county's lagged value is then the weighted sum of its neighbours' values.

This is the most intuitive way to summarise a neighbourhood, but it has a real drawback for counties that are found near the edges of the study area. A county with only one or two neighbours has each of those neighbours carrying a much larger share of the total weight, so its lagged value depends heavily on just a handful of counties. This can potentially over- or under-estimate the true spatial pattern in that area.

Row standardisation is used here for simplicity. A more robust alternative, **style = "B"**, drops the equal-share standardisation and instead gives every neighbour a raw weight of 1, producing a neighbourhood total rather than an average.

```{r}
rswm_q <- nb2listw(wm_q, style = "W", zero.policy = TRUE)
rswm_q
```

::: {.callout-warning}
Setting **zero.policy = TRUE** permits counties with no neighbours to be carried through as empty sets rather than raising an error. This should be used with caution - it converts a condition that would otherwise stop the analysis into a silent one, and the user may not be aware that some units in their dataset have no neighbours at all. 

Setting **zero.policy = FALSE** would cause the function to return an error whenever such units exist, surfacing the issue immediately rather than letting it pass unnoticed.
:::

The weights attached to the 10th county's neighbours:

```{r}
rswm_q$weights[10]
```

County 10 has 8 neighbours, each assigned a weight of 1/8 = 0.125. This means that when the spatial lag is computed, each neighbour's GDPPC will be multiplied by 0.125 before being summed.

An inverse distance weights list is constructed the same way, passing `ids` through the `glist` argument:

```{r}
idw_b <- nb2listw(wm_q, glist = ids, style = "B", zero.policy = TRUE)
idw_b
```

::: {callout-note}
Under style = "B", the values supplied through `glist` are used exactly as given, with no rescaling applied.
:::

The weights attached to the first county's neighbours:

```{r}
idw_b$weights[1]
```

::: {.callout-note}
`idw_b` is not row-standardised. The above output sums to approximately 0.112 (not 1), and `S0` in the weights constants summary is 8.786867 rather than 88, the number of regions. Under row standardisation (style = "W"), every row is rescaled to sum to exactly 1, and `S0` equals `n`, as confirmed earlier for `rswm_q`.

`idw_b` still reflects the same distance-decay relationship, where closer neighbours carry larger weights, but its absolute magnitude depends on the units of distance. Hence, the lagged values are not directly comparable across counties with different neighbour structures.
:::

```{r}
summary(unlist(idw_b$weights))
```

## 9 Application of Spatial Weight Matrices

This section will cover four different spatial lagged variables:

| Variant | Weighting | Includes self |
|---|---|---|
| Spatial lag | Row-standardised | No |
| Spatial lag as a sum | Binary | No |
| Spatial window average | Row-standardised | Yes |
| Spatial window sum | Binary | Yes |

### 9.1 Spatial lag with row-standardised weights

The average neighbouring GDPPC value is computed for each county using `lag.listw()`. These values are often referred to as **spatially lagged values**.

```{r}
GDPPC.lag <- lag.listw(rswm_q, hunan$GDPPC)
GDPPC.lag
```

As an illustration, the GDPPC of the 1st county's neighbours is retrieved below.

```{r}
nb1 <- wm_q[[1]]
nb1 <- hunan$GDPPC[nb1]
nb1
```

::: {.callout-note}
**Interpretation:**

Under row-standardised weights, each of the five neighbours receives an equal weight of 1/5 = 0.2. The spatial lag is computed by multiplying each neighbour's GDPPC by that weight, then summing the results:

(20,981 × 0.2) + (34,592 × 0.2) + (24,473 × 0.2) + (21,311 × 0.2) + (22,879 × 0.2) = **24,847.20**

This value (24,847.20) represents the 1st county's spatial lag of GDPPC.
:::

The spatially lagged GDPPC values, `GDPPC.lag`, computed above for all 88 counties, are appended to `hunan` and mapped alongside the raw indicator:

```{r}
lag.list <- list(hunan$NAME_3, GDPPC.lag)
lag.res <- as.data.frame(lag.list)
colnames(lag.res) <- c("NAME_3", "lag GDPPC")
hunan <- left_join(hunan, lag.res, by = "NAME_3")
```

The following output shows the average neighbouring GDPPC values for each county.

```{r}
head(hunan)
```

Both GDPPC and spatially lagged GDPPC are plotted below for comparison:

```{r}
gdppc <- qtm(hunan, "GDPPC")
lag_gdppc <- qtm(hunan, "lag GDPPC")
tmap_arrange(gdppc, lag_gdppc, asp = 1, ncol = 2)
```

::: {.callout-note}
**Interpretation:**

The lagged map confirms the distinction drawn earlier between a hotspot and a spatial outlier. The Changsha cluster remains dark, since Changsha, Wangcheng, Liuyang, and Ningxiang are mutually adjacent and they are all individually wealthy, so each one pulls the others' neighbourhood averages upward. In constrast, Zixing and Lengshuijiang fade out entirely. Even though both counties have relatively high GDPPC, they are surrounded by poorer neighbours, hence their neighbourhood averages sit below their own GDPPC.
:::

### 9.2 Spatial lag as a sum of neighbouring values

Spatial lag as a sum of neighbouring values is computed by assigning binary weights instead of row-standardised ones. This requires returning to the neighbours list, applying a function that assigns each neighbour a weight, then passing those weights into *nb2listw()* through the `glist` argument, so they are used exactly as supplied rather than derived automatically from neighbour count.

The first step assigns a value of 1 to every neighbour. This is done with *lapply()*, the same function used throughout this document to operate across a neighbours list: it applies a given function to each element of the list in turn, one county's neighbour vector at a time.

```{r}
b_weights <- lapply(wm_q, function(x) 0 * x + 1)
b_weights2 <- nb2listw(wm_q, glist = b_weights, style = "B")
b_weights2
```

With the proper weights assigned, `lag.listw()` computes the neighbourhood sum in the same way as before, using `b_weights2` in place of the row-standardised weights:

```{r}
lag_sum <- list(hunan$NAME_3, lag.listw(b_weights2, hunan$GDPPC))
lag.res <- as.data.frame(lag_sum)
colnames(lag.res) <- c("NAME_3", "lag_sum GDPPC")
```

```{r}
lag_sum
```

::: {.callout-note}
**Interpretation:**

Spatial lag as a sum of neighbouring values is the total GDPPC of all a county's neighbours added together, rather than their average. Every neighbour contributes its full value to that total, and therefore a county's `lag_sum` grows both with how wealthy its neighbours are and with how many neighbours it has.

This is different from `GDPPC.lag`, which divides that same total by the neighbour count to produce an average. A simple formula connects the two: `lag_sum` equals `GDPPC.lag` multiplied by the number of neighbours. Anxiang's lag_sum of 124,236, divided by its 5 neighbours, gives 24,847.20, exactly the same as GDPPC.lag value for this county.
:::

The lag_sum GDPPC field is appended into the `hunan` sf data frame:

```{r}
hunan <- left_join(hunan, lag.res, by = "NAME_3")
```

Both GDPPC and spatial lag sum GDPPC are plotted for comparison:

```{r}
lag_sum_gdppc <- qtm(hunan, "lag_sum GDPPC")
tmap_arrange(gdppc, lag_sum_gdppc, asp = 1, ncol = 2)
```

::: {.callout-note}
**Interpretation:**

From the result, the county with the darkest shade is Xiangtan (County 86), noticeably darker than its neighbours in the cluster. This is interesting because Xiangtan's own GDPPC is only 27,060, nowhere near the top of the province. Its lag_sum is highest not because Xiangtan itself is wealthy, but it happens to border nine counties at once, including Changsha (88,656), Wangcheng (70,666), and Ningxiang (62,202), three of the province's wealthiest counties, all landing as direct neighbours of one modestly-ranked county.
:::

### 9.3 Spatial window average

The spatial window average uses row-standardised weights, in the same way as the first spatial lag computed earlier, but the neighbourhood now includes the county's own value alongside its neighbours'. This requires returning to the neighbours list and adding the diagonal element before the weights are assigned, using `include.self()` of **spdep**:

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Compared to `wm_q`, the number of nonzero links increases from 448 to 536 (88 more), since each county has gained itself as an additional neighbour. The average number of links also rises correspondingly, from 5.09 to 6.09.

The change is directly visible at the county level. Anxiang's neighbour list, which held 5 entries under `wm_q`, now holds 6.

```{r}
wm_qs[[1]]
```

Weights are then derived with `nb2listw()`:

```{r}
wm_qs <- nb2listw(wm_qs)
wm_qs
```

Next, the lag variable is created from the weight structure and GDPPC variable.

```{r}
lag_w_avg_gdppc <- lag.listw(wm_qs, hunan$GDPPC)
lag_w_avg_gdppc
```

::: {.callout-note}
**Interpretation:**

Anxiang's window average is 24,650.50, close to the first spatial lag (`GDPPC.lag`) of 24,847.20 computed earlier. The difference is due to the addition of Anxiang's own GDPPC (23,667) into the average: (20,981 + 34,592 + 24,473 + 21,311 + 22,879 + 23,667) / 6 = 24,650.50. Since Anxiang's own value is slightly below its neighbourhood average, including it pulls the window average down slightly. 

Whether the window average moves up or down relative to the plain lag depends on whether a county's own value sits above or below its neighbours' average, and by how much.
:::

The lag_window_avg GDPPC field is appended into the `hunan` sf data frame:

```{r}
lag.list.wm_qs <- list(hunan$NAME_3, lag_w_avg_gdppc)
lag_wm_qs.res <- as.data.frame(lag.list.wm_qs)
colnames(lag_wm_qs.res) <- c("NAME_3", "lag_window_avg GDPPC")
hunan <- left_join(hunan, lag_wm_qs.res, by = "NAME_3")
```

To compare the values of lag GDPPC and spatial window average, `kable()` of **knitr** package is used.

```{r}
hunan %>%
  select("County", 
         "lag GDPPC", 
         "lag_window_avg GDPPC") %>%
  kable()
```

`qtm()` of **tmap** is used to plot `lag_gdppc` and `w_avg_gdppc` for comparison:

```{r}
w_avg_gdppc <- qtm(hunan, "lag_window_avg GDPPC")
tmap_arrange(lag_gdppc, w_avg_gdppc, asp = 1, ncol = 2)
```

::: {.callout-note}
**Interpretation:**

The two maps show different counties with the darker shade. Under plain lag, Pingjiang ranks highest at 55,157.75, since its four neighbours are, on average, very wealthy. Under window average, Pingjiang drops to 47,576.60. This is because Pingjiang's own GDPPC is only 17,252, well below its neighbours' average, so adding its own lower value to the average pulls the result down, now that five values are being averaged instead of four. This highlights the effect that fewer neighbours would mean that a county's own value carries more significant weight than it would with more neighbours.
:::

### 9.4 Spatial window sum

The spatial window sum is the counter part of the window average, but without using row-standardized weights.

To add the diagonal element to the neighbour list, `include.self()` of **spdep** will be used.

```{r}
wm_qs <- include.self(wm_q)
wm_qs
```

Next, binary weights are assigned to the neighbour structure, giving every entry, including the county's own added position, a weight of 1:

```{r}
b_weights <- lapply(wm_qs, function(x) 0 * x + 1)
b_weights[1]
```

Position 1 now returns six 1s instead of five, one more than under `wm_q`, confirming the diagonal element has been included in the weighting.

```{r}
b_weights2 <- nb2listw(wm_qs, glist = b_weights, style = "B")
b_weights2
```

The lag variable is computed the same way as `lag_sum` earlier, using the new weight structure:

```{r}
w_sum_gdppc <- lag.listw(b_weights2, hunan$GDPPC)
w_sum_gdppc
```

The w_sum GDPPC field is appended into the `hunan` sf data frame:

```{r}
w_sum_gdppc_list <- list(hunan$NAME_3, w_sum_gdppc)
w_sum_gdppc.res <- as.data.frame(w_sum_gdppc_list)
colnames(w_sum_gdppc.res) <- c("NAME_3", "w_sum GDPPC")
hunan <- left_join(hunan, w_sum_gdppc.res, by = "NAME_3")
```

To compare the values of lag sum GDPPC and spatial window sum, `kable()` of **knitr** package is used.

```{r}
hunan %>%
  select(County, `lag_sum GDPPC`, `w_sum GDPPC`) %>%
  kable()
```

`qtm()` of **tmap** is used to plot `lag_sum_gdppc` and `w_sum_gdppc` for comparison:

```{r}
w_sum_gdppc_map <- qtm(hunan, "w_sum GDPPC")
tmap_arrange(lag_sum_gdppc, w_sum_gdppc_map, asp = 1, ncol = 2)
```

::: {.callout-note}
**Interpretation:**

The two maps look similar as Xiangtan dominates both, rising from 404,456 under `lag_sum` to 431,516 under `w_sum`.

The window sum possesses the same limitation as lag sum: it mixes neighbourhood size and neighbourhood wealth into a single total. Adding the county's own value does not fix this. It just adds a third factor into the mix. A county can rank highly for any of these reasons: it has many neighbours, one of its neighbours is very wealthy, or the county itself is wealthy. The total alone does not show which of these is actually happening.
:::

## 10 Reference

- [Creating Neighbours using sf objects](https://cran.r-project.org/web/packages/spdep/vignettes/nb_sf.html)