---
title: "In-class Exercise 6"
date: "October 03, 2026"
date-modified: today
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1 Installing and Loading the R Packages

Six R packages are required for this in-class exercise:

```{r}
pacman::p_load(sf, sfdep, tmap, 
               plotly, tidyverse, 
               Kendall)
```

## 2 The Data

For the purpose of this exercise, the Hunan data sets will be used. There are two data sets

- Hunan, a geospatial data set in ESRI shapefile format, and
- Hunan_GDPPC, an attribute data set in csv format.

### 2.1 Importing geospatial data

In the code chunk below, `st_read()` of **sf** package is used to import Hunan shapefile into R.

```{r}
hunan <- st_read(
  dsn = "data/geospatial", 
  layer = "Hunan") %>% 
  st_transform(crs = 32650)
```

### 2.2 Importing attribute table

In the code chunk below, 'read_csv()' of readr is used to import `Hunan_GDPPC.csv` into R.

```{r}
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

## 3 Creating a Time Series Cube

In the code chunk below, `spacetime()` of **sfdep** is used to create an spatio-temporal cube.

```{r}
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, `is_spacetime_cube()` of **sfdep** package will be used to verify if `GDPPC_st` is a space-time cube object.

```{r}
is_spacetime_cube(GDPPC_st)
```

The **TRUE** return confirms that `GDPPC_st` object is a time-space cube.

## 4 Computing Gi* (Part 1)

### 4.1 Deriving the spatial weights

The code chunk below will be used to identify neighbors and to derive an inverse distance weights.

```{r}
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(
    st_contiguity(geometry)),
    wt = st_inverse_distance(nb, 
                             geometry, 
                             scale = 1,
                             alpha = 1),
    .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

## 5 Computing Gi* (Part 2)

### 5.1 Deriving the spatial weights

We can use the new columns to manually calculate the local Gi* for each location. This can be performed by grouping by `Year` and using `local_gstar_perm()` of **sfdep** package. After which, we use `unnest()` to unnest gi_star column of the newly created `gi_starts` data.frame.

```{r}
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

## 6 Mann-Kendall Test

A monotonic series or function is one that only increases (or decreases) and never changes direction. So long as the function either stays flat or continues to increase, it is monotonic.

Null Hypothesis (H0): No monotonic trend

Alternative Hypothesis (H1): Monotonic trend is present

**Interpretation:**

- Reject the null-hypothesis null if the p-value is smaller than the alpha value (i.e. 1-confident level)
- Tau ranges between -1 and 1 where:
  - -1 is a perfectly decreasing series, and
  - 1 is a perfectly increasing series.

### 6.1 Mann-Kendall Test on Gi

With these Gi* measures we can evaluate each location for a trend using the Mann-Kendall test. 

The code chunk below uses Changsha county.

```{r}
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

Next, the result can be plotted using **ggplot2** functions.

```{r}
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

### 6.2 Interactive Mann-Kendall Plot

We can also create an interactive plot by using `ggplotly()` of plotly package.

```{r}
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

### 6.3 Printing Mann-Kendall test report

```{r}
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

`sl` is the p-value. With reference to the results, since p-value is lower than 0.05, we will reject the null hypothesis and infer that a slight upward trend.

### 6.4 Mann-Kendall test data.frame

We can replicate this for each location by using `group_by()` of **dplyr** package.

```{r}
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
head(ehsa)
```

We can also sort to show significant emerging hot/cold spots:

```{r}
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:10)
head(emerging)
```

## 7 Performing Emerging Hotspot Analysis

Lastly, we will perform EHSA analysis by using `emerging_hotspot_analysis()` of **sfdep** package. It takes a spacetime object x (i.e. GDPPC_st), and the quoted name of the variable of interest (i.e. GDPPC) for .var argument. The k argument is used to specify the number of time lags which is set to 1 by default. nsim map numbers of simulation to be performed.

```{r}
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

### 7.1 Visualising the distribution of EHSA classes

ggplot2 functions is used to reveal the distribution of EHSA classes as a bar chart.

```{r}
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

The figure above shows that sporadic cold spots class has the high numbers of county.

### 7.2 Visualising EHSA

In this section, we will visualise the geographic distribution EHSA classes. However, before that, we will need to join both `hunan` and `ehsa` together.

```{r}
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

Next, tmap functions will be used to plot a categorical choropleth map by using the code chunk below.

```{r}
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(fill_alpha = 0.4)
```


