---
title: "Hands-on Exercise 1(A): Geospatial Data Science with R"
---

Geospatial Data Science is a process of importing, wrangling, integrating, and processing geographically referenced data sets.

In this hands-on exercise, we will learn how to perform geospatial data science tasks in R by using **sf** package.

## 1.1 Learning Outcome

This exercise aims to cover the following:

- installing and loading [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://tidyverse.org/) packages into R environment,
- importing geospatial data by using appropriate functions of **sf** package,
- importing aspatial data by using appropriate function of **readr** package,
- exploring the content of simple feature data frame by using appropriate **Base R** and **sf** functions,
- assigning or transforming coordinate systems by using using appropriate **sf** functions,
- converting an aspatial data into a sf data frame by using appropriate function of **sf** package,
- performing geoprocessing tasks by using appropriate functions of **sf** package,
- performing data wrangling tasks by using appropriate functions of **dplyr** package and
- performing Exploratory Data Analysis (EDA) by using appropriate functions from **ggplot2** package.

## 1.2 Data Acquisition

In this hands-on exercise, we will be using the following data sets:

| Dataset | Source | Description |
|------------------------|------------------------|------------------------|
| **Master Plan 2014 Subzone Boundary (No Sea)** | [data.gov.sg](https://data.gov.sg/) | Geospatial boundaries for Singapore's planning subzones |
| **Pre-Schools Location** | [data.gov.sg](https://data.gov.sg/) | Location data for pre-schools in Singapore |
| **Cycling Path** | [LTA DataMall](https://datamall.lta.gov.sg/content/datamall/en.html) | Geospatial data for cycling paths in Singapore |
| **Singapore Airbnb Listings** | [Inside Airbnb](http://insideairbnb.com/get-the-data.html) | Listing data for Airbnb properties in Singapore |

::: callout-note
The Master Plan 2014 Subzone Boundary and Pre-Schools Location data sets were downloaded in **GeoJSON** format, as data.gov.sg no longer offers these particular data sets in Shapefile or KML format respectively.
:::

### 1.2.1 Extracting the geospatial data sets

To get started, create a new folder called *Hands-on_Ex* in a working directory.

Next, create a new sub-folder called **Hands-on_Ex01** in the newly created *Hands-on_Ex* folder.

In the `Hands-on_Ex01` folder, create a sub-folder called `data`. Then, inside the `data` sub-folder, create two sub-folders and name them `geospatial` and `aspatial` respectively.

Place `Master Plan 2014 Subzone Boundary (No Sea)`, `Pre-Schools Location` and `Cycling Path` zipped files into `geospatial` sub-folder and unzipped them. Copy the unzipped files from their respective sub-folders and place them inside `geospatial` sub-folder.

### 1.2.2 Extracting the aspatial data set

Extract the downloaded `listings.csv.gz` file, and move the resulting `listings.csv` into the `aspatial` sub-folder.

## 1.3 Getting Started

In this hands-on exercise, two R packages will be used. They are:

- **sf** for importing, managing, and processing geospatial data, and
- **tidyverse** for performing data science tasks such as importing, wrangling and visualising data.

Use the code chunk below to load the necessary R packages into R.

```{r}
pacman::p_load(sf, tidyverse)
```

## 1.4 Importing Geospatial Data

In this section, three geospatial datasets are imported into R using `st_read()` of the **sf** package:

- `MP14_SUBZONE_NO_SEA_PL`, a polygon feature layer representing Singapore's planning subzones (Master Plan 2014, No Sea variant)
- `CyclingPathGazette`, a line feature layer representing Singapore's cycling paths
- `PreSchoolsLocation`, a point feature layer representing pre-school locations

### 1.4.1 Importing polyline feature data in GeoJSON format (Master Plan 2024 Subzone Boundary)

The code chunk below uses *st_read()* function of **sf** package to import `MP14_SUBZONE_NO_SEA_PL` into R as a polygon feature data frame. Since the input data is in **GeoJSON** format, only a single `dsn` (data source name) argument is needed to define the file path.

```{r}
mpsz = st_read(dsn = "data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

The message above reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 13 fields in `mpsz` simple feature data frame. `mpsz` is in **wgs84** geodetic coordinates system — a geographic coordinate system using latitude/longitude in degrees. The bounding box provides the x extent and y extent of the data.

### 1.4.2 Importing polyline feature data in shapefile format (Cycling Path)

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as a line feature data frame.

```{r}
cyclingpath = st_read(dsn = "data/geospatial",
                       layer = "CyclingPathGazette")
```

The message above reveals that the geospatial objects are multilinestring features. There are a total of 4,830 multilinestring features and 7 fields in `cyclingpath` simple feature data frame. This data set is in **svy21** projected coordinates system, where its coordinates are expressed in metres.

### 1.4.3 Importing GIS data in GeoJSON format (Pre-school Locations)

The `PreSchoolsLocation` is in **GeoJSON** format. The code chunk below will be used to import `PreSchoolsLocation` into R.

```{r}
preschool = st_read(dsn = "data/geospatial/PreSchoolsLocation.geojson") %>%
  st_transform(crs = 3414)
```

::: callout-note
**Code Explanation:** Read the `preschoolslocation` GeoJSON file into R, then take that result and convert its coordinates from WGS84 (degrees) into SVY21 (metres), and store the final output in a variable called preschool.
:::

The message above reveals that `preschool` is a point feature data frame. There are a total of 2290 features and 2 fields. Similar to `mpsz`, `preschool` is also in **wgs84** coordinates system.

## 1.5 Checking the Content of A Simple Feature Data Frame

### 1.5.1 Working with st_geometry()

The column in the sf data.frame that contains the geometries is a list of class sfc. We can retrieve the geometry list-column in this case by mpsz$geom or mpsz[[1]], but the more general way uses [st_geometry()](https://r-spatial.github.io/sf/reference/st_geometry.html) as shown in the code chunk below.

```{r}
st_geometry(mpsz, n=0)
```

