---
title: "Hands-on Exercise 1(A): Geospatial Data Science with R"
---

Geospatial Data Science is a process of importing, wrangling, integrating, and processing geographically referenced data sets.

In this hands-on exercise, we will learn how to perform geospatial data science tasks in R by using **sf** package.

## 1.1 Learning Outcome

This exercise aims to cover the following:

- installing and loading [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://tidyverse.org/) packages into R environment,
- importing geospatial data by using appropriate functions of **sf** package,
- importing aspatial data by using appropriate function of **readr** package,
- exploring the content of simple feature data frame by using appropriate **Base R** and **sf** functions,
- assigning or transforming coordinate systems by using using appropriate **sf** functions,
- converting an aspatial data into a sf data frame by using appropriate function of **sf** package,
- performing geoprocessing tasks by using appropriate functions of **sf** package,
- performing data wrangling tasks by using appropriate functions of **dplyr** package and
- performing Exploratory Data Analysis (EDA) by using appropriate functions from **ggplot2** package.

## 1.2 Data Acquisition

In this hands-on exercise, we will be using the following data sets:

| Dataset | Source | Description |
|------------------------|------------------------|------------------------|
| **Master Plan 2014 Subzone Boundary (No Sea)** | [data.gov.sg](https://data.gov.sg/) | Geospatial boundaries for Singapore's planning subzones |
| **Pre-Schools Location** | [data.gov.sg](https://data.gov.sg/) | Location data for pre-schools in Singapore |
| **Cycling Path** | [LTA DataMall](https://datamall.lta.gov.sg/content/datamall/en.html) | Geospatial data for cycling paths in Singapore |
| **Singapore Airbnb Listings** | [Inside Airbnb](http://insideairbnb.com/get-the-data.html) | Listing data for Airbnb properties in Singapore |

::: callout-note
The Master Plan 2014 Subzone Boundary and Pre-Schools Location data sets were downloaded in **GeoJSON** format, as data.gov.sg no longer offers these particular data sets in Shapefile or KML format respectively.
:::

### 1.2.1 Extracting the geospatial data sets

To get started, create a new folder called *Hands-on_Ex* in a working directory.

Next, create a new sub-folder called **Hands-on_Ex01** in the newly created *Hands-on_Ex* folder.

In the `Hands-on_Ex01` folder, create a sub-folder called `data`. Then, inside the `data` sub-folder, create two sub-folders and name them `geospatial` and `aspatial` respectively.

Place `Master Plan 2014 Subzone Boundary (No Sea)`, `Pre-Schools Location` and `Cycling Path` zipped files into `geospatial` sub-folder and unzipped them. Copy the unzipped files from their respective sub-folders and place them inside `geospatial` sub-folder.

### 1.2.2 Extracting the aspatial data set

Extract the downloaded `listings.csv.gz` file, and move the resulting `listings.csv` into the `aspatial` sub-folder.

## 1.3 Getting Started

In this hands-on exercise, two R packages will be used. They are:

- **sf** for importing, managing, and processing geospatial data, and
- **tidyverse** for performing data science tasks such as importing, wrangling and visualising data.

Use the code chunk below to load the necessary R packages into R.

```{r}
pacman::p_load(sf, tidyverse)
```

## 1.4 Importing Geospatial Data

In this section, three geospatial datasets are imported into R using [st_read()](https://r-spatial.github.io/sf/reference/st_read.html) of the **sf** package:

- `MP14_SUBZONE_NO_SEA_PL`, a polygon feature layer representing Singapore's planning subzones (Master Plan 2014, No Sea variant)
- `CyclingPathGazette`, a line feature layer representing Singapore's cycling paths
- `PreSchoolsLocation`, a point feature layer representing pre-school locations

### 1.4.1 Importing polyline feature data in GeoJSON format (Master Plan 2024 Subzone Boundary)

The code chunk below uses *st_read()* function of **sf** package to import `MP14_SUBZONE_NO_SEA_PL` into R as a polygon feature data frame. Since the input data is in **GeoJSON** format, only a single `dsn` (data source name) argument is needed to define the file path.

```{r}
mpsz = st_read(dsn = "data/geospatial/MasterPlan2014SubzoneBoundaryNoSea.geojson")
```

The message above reveals that the geospatial objects are multipolygon features. There are a total of 323 multipolygon features and 13 fields in `mpsz` simple feature data frame. `mpsz` is in **wgs84** geodetic coordinates system — a geographic coordinate system using latitude/longitude in degrees. The bounding box provides the x extent and y extent of the data.

### 1.4.2 Importing polyline feature data in shapefile format (Cycling Path)

The code chunk below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as a line feature data frame.

```{r}
cyclingpath = st_read(dsn = "data/geospatial",
                       layer = "CyclingPathGazette")
```

The message above reveals that the geospatial objects are multilinestring features. There are a total of 4,830 multilinestring features and 7 fields in `cyclingpath` simple feature data frame. This data set is in **svy21** projected coordinates system, where its coordinates are expressed in metres.

### 1.4.3 Importing GIS data in GeoJSON format (Pre-school Locations)

The `PreSchoolsLocation` is in **GeoJSON** format. The code chunk below will be used to import `PreSchoolsLocation` into R.

```{r}
preschool = st_read(dsn = "data/geospatial/PreSchoolsLocation.geojson")
```

The message above reveals that `preschool` is a point feature data frame. There are a total of 2290 features and 2 fields. Similar to `mpsz`, `preschool` is in **wgs84** coordinates system.

## 1.5 Checking the Content of A Simple Feature Data Frame

### 1.5.1 Working with st_geometry()

The column in the sf data.frame that contains the geometries is a list of class `sfc`. We can retrieve the geometry list-column in this case by mpsz$geom or mpsz[[1]], but the more general way uses [st_geometry()](https://r-spatial.github.io/sf/reference/st_geometry.html) as shown in the code chunk below.

```{r}
st_geometry(mpsz)
```

### 1.5.2 Working with glimpse()

Beyond the basic geospatial feature information, an sf data frame also holds associated attribute information. The code chunk below uses [glimpse()](https://dplyr.tidyverse.org/reference/glimpse.html) function of **dplyr** package to inspect this attribute data.

```{r}
glimpse(mpsz)
```

`glimpse()` report reveals the data type of each fields. For example, `SHAPE_1.AREA` and `SHAPE_1.LEN` fields are in **double-precision values**.

### 1.5.3 Working with head()

While `glimpse()` gives a quick overview of all fields, sometimes it's useful to view the complete information of individual features in full. This can be done using [head()](https://www.rdocumentation.org/packages/utils/versions/3.6.2/topics/head) function of Base R, as shown in the code chunk below.

```{r}
head(mpsz, n=5)
```

::: callout-note
One of the useful arguments of head() is that it allows the user to select the number of records to display (i.e. the n argument).
:::

## 1.6 Plotting the Geospatial Data

Beyond reviewing feature information, visualising the geospatial features is equally important in geospatial data science. The code chunk below uses `plot()` function of Base R Graphics to do so.

```{r}
#| warning: false
plot(mpsz)
```

::: callout-note
The default `plot()` of an sf object is a multi-plot of all attributes, up to a reasonable maximum as shown above. 
:::

We can also plot only the geometry by using the code chunk below.

```{r}
plot(st_geometry(mpsz))
```

Alternatively, a specific attribute of the `sf` object can also be plotted on its own, as shown in the code chunk below.

```{r}
plot(mpsz["PLN_AREA_N"])
```

::: {.callout-tip}
`plot()` is appropriate for a quick, exploratory look at geospatial data. For higher cartographic-quality maps, other R packages such as **tmap** should be used.
:::

Next, to see how two data sets relate spatially, the `preschool` layer can be plotted on top of the `mpsz` layer, as shown in the code chunk below.

```{r}
#| warning: false
plot(st_geometry(mpsz))
plot(preschool, add = TRUE)
```

## 1.7 Working with Projection

Map projection is an important property of a geospatial data. In order to perform geoprocessing using two geospatial data, we need to ensure that both geospatial data are projected using similar coordinate system.

This section covers how to convert a simple feature data frame from one coordinate system to another — a process known as **projection transformation**.

### 1.7.1 Transforming the projection of mpsz from wgs84 to svy21

In geospatial analytics, it is common to transform data from a geographic coordinate system to a projected coordinate system, since geographic coordinate systems are not appropriate when analysis requires distance and/or area measurements.

The code chunk below performs the projection transformation on `mpsz`.

```{r}
mpsz <- st_transform(mpsz,
                     crs = 3414)
```

```{r}
st_geometry(mpsz)
```

::: callout-note
The output confirms that `mpsz` is in svy21 projected coordinate system. Furthermore, referring to the Bounding box, the values are greater than the 0–360 decimal degree range commonly used by most of the geographic coordinate systems.
:::

### 1.7.2 Transforming the projection of preschool from wgs84 to svy21

The code chunk below performs the projection transformation on `preschool`.

```{r}
preschool <- st_transform(preschool,
                          crs = 3414)
```

```{r}
st_geometry(preschool)
```

::: callout-note
The output confirms that `preschool` is now in svy21 projected coordinate system.
:::

With both `mpsz` and `preschool` now in the same svy21 coordinate system, the `preschool` layer can be plotted on top of the `mpsz` layer once again.

```{r}
#| warning: false
plot(st_geometry(mpsz))
plot(st_geometry(preschool), add = TRUE)
```

## 1.8 Importing and Converting An Aspatial Data

In practice, it is not unusual to come across data such as `listings` from Inside Airbnb. This kind of data is referred to as aspatial data — it is not geospatial data, but among the data fields, there are two fields that capture the x- and y-coordinates of the data points.

This section covers how to import aspatial data into R and save it as a tibble data frame, before converting it into a simple feature data frame.

### 1.8.1 Importing the aspatial data

Since the `listings` data set is in csv file format, [read_csv()](https://readr.tidyverse.org/reference/read_delim.html) function of **readr** package is used to import `listings.csv`, as shown in the code chunk below. The output R object is called `listings`, and it is a [tibble data frame](https://r4ds.had.co.nz/tibbles.html).

```{r}
#| message: false
listings <- read_csv("data/aspatial/listings.csv")
```

After importing the data file into R, it is important to examine if the file has been imported correctly.

The code chunk below uses `list()` of Base R, instead of `glimpse()`, to examine the imported data.

```{r}
list(listings)
```

The output reveals that the `listings` tibble data frame consists of **3,247 rows** and **19 columns**. Two useful fields to be used in the next phase are `latitude` and `longitude`, which are in decimal degree format. As a best guess, this data is assumed to be in the **wgs84** geographic coordinate system.

### 1.8.2 Creating a simple feature data frame from an aspatial data frame

The code chunk below converts listing data frame into a simple feature data frame by using [st_as_sf()](https://r-spatial.github.io/sf/reference/st_as_sf.html) of **sf** packages.

```{r}
listings_sf <- st_as_sf(listings, 
                       coords = c("longitude", "latitude"),
                       crs=4326) %>%
  st_transform(crs = 3414)
```

A few things to note about the arguments above:

- coords argument requires you to provide the column name of the x-coordinates first then followed by the column name of the y-coordinates.
- crs argument requires you to provide the coordinates system in epsg format. [EPSG](https://epsg.io/4326): 4326 is wgs84 Geographic Coordinate System and [EPSG](https://epsg.io/3414): 3414 is Singapore SVY21 Projected Coordinate System. You can search for other country’s epsg code by referring to [epsg.io](https://epsg.io/).
- %>% is used to nest `st_transform()` to transform the newly created simple feature data frame into svy21 projected coordinates system.

The content of this newly created simple feature data frame can be examined below.

```{r}
glimpse(listings_sf)
```

The table above shows the content of `listings_sf`. A new `geometry` column has been added to the data frame, while the original `longitude` and `latitude` columns have been dropped.

The map below shows the listings_sf layer plotted on top of mpsz, with each point representing an Airbnb listing across Singapore.

```{r}
#| warning: false
plot(st_geometry(mpsz))
plot(st_geometry(listings_sf), add = TRUE)
```

## 1.9 Geoprocessing with sf package

