---
title: "Hands-on Exercise 6: Geographical Segmentation with Spatially Constrained Clustering Techniques"
date: "September 28, 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1 Overview

In this hands-on exercise, the focus will be on how to delineate homogeneous region by using geographically referenced multivariate data. There are two major analysis, namely:

- hierarchical cluster analysis; and
- spatially constrained cluster analysis.

### 1.1 Learning outcome

By the end of this exercise, the following are covered:

- converting GIS polygon data into an **sf** simple feature data frame;
- converting the simple feature data frame into a SpatialPolygonsDataFrame object by using `as_Spatial()` of the **sf** package;
- performing cluster analysis by using `hclust()` of base R;
- performing spatially constrained cluster analysis by using `skater()` of **spdep**; and
- visualising the analysis output by using the **ggplot2** and **tmap** packages.

## 2 Analytical Question

In geobusiness and spatial policy, it is common practice to delineate a market or planning area into homogeneous regions by using multivariate data. In this exercise, the focus is to delineate [Shan State, Myanmar](https://en.wikipedia.org/wiki/Shan_State) into homogeneous regions by using six Information and Communication Technology (ICT) measures: radio, television, land line phone, mobile phone, computer and internet at home.

## 3 The Data

Two data sets are used in this study:

| Data set | Format | Description |
|------|------|------|
| myanmar_township_boundaries | ESRI shapefile | Myanmar township boundary information |
| Shan-ICT.csv | CSV | Extract of the [2014 Population and Housing Census of Myanmar](https://myanmar.unfpa.org/en/publications/2014-population-and-housing-census-myanmar-data-sheet) at the township level |

Both data sets were downloaded from [Myanmar Information Management Unit (MIMU)](https://themimu.info/).

## 4 Installing and Loading R packages

The R packages needed for this exercise are:

- Spatial data handling: **sf** and **spdep**;
- Attribute data handling: **tidyverse**, especially **readr**, **ggplot2** and **dplyr**;
- Choropleth mapping: **tmap**;
- Multivariate data visualisation and analysis: **corrplot**, **psych**, **ggpubr**, **heatmaply** and **GGally**; and
- Cluster analysis: **cluster**, **factoextra**, **NbClust** and **ClustGeo**.

```{r}
pacman::p_load(sf, spdep, tidyverse, tmap,
               corrplot, psych, ggpubr, heatmaply, GGally,
               cluster, factoextra, NbClust, ClustGeo)
```

## 5 Data Import and Preparation

### 5.1 Importing geospatial data

The Myanmar township boundary data is in ESRI shapefile format and is imported with `st_read()` of **sf**.

```{r}
myanmar_sf <- st_read(dsn = "data/geospatial",
                      layer = "myanmar_township_boundaries")
```

Shan State is split into three labels in the `ST` field.

```{r}
myanmar_sf %>%
  st_drop_geometry() %>%
  count(ST, sort = TRUE)
```

The three Shan labels are: Shan (North), Shan (South) and Shan (East).

```{r}
shan_sf <- myanmar_sf %>%
  filter(ST %in% c("Shan (North)", "Shan (South)", "Shan (East)")) %>%
  select(c(2:7))
```

The imported township boundary object is called `shan_sf` and is saved in simple feature data frame format.

```{r}
shan_sf
```

Since `shan_sf` conforms to the tidy framework, `glimpse()` can also be used to reveal the data type of each field:

```{r}
glimpse(shan_sf)
```

### 5.2 Importing aspatial data

The CSV file is imported with `read_csv()` of **readr**:

```{r}
ict <- read_csv("data/aspatial/Shan-ICT.csv")
```

The imported ICT variables are extracted from the 2014 Population and Housing Census of Myanmar. It is saved in R’s tibble data.frame format.

The summary statistics are revealed with the code chunk below:

```{r}
summary(ict)
```

There are 11 fields and 55 observations in the tibble, with no missing values.

### 5.3 Deriving new variables using dplyr package

The unit of measurement of the ICT fields is the number of households. Using these values directly would be biased by the total number of households. In general, townships with more households will also have more households owning a radio, a television and so on.

To overcome this, the penetration rate of each ICT variable is derived as households per 1,000 households, using the code chunk below:

```{r}
ict_derived <- ict %>%
  rename(DT_PCODE = `District Pcode`, DT = `District Name`,
         TS_PCODE = `Township Pcode`, TS = `Township Name`,
         TT_HOUSEHOLDS = `Total households`,
         RADIO = Radio, TV = Television,
         LLPHONE = `Land line phone`, MPHONE = `Mobile phone`,
         COMPUTER = Computer, INTERNET = `Internet at home`) %>%
  mutate(RADIO_PR = RADIO / TT_HOUSEHOLDS * 1000,
         TV_PR = TV / TT_HOUSEHOLDS * 1000,
         LLPHONE_PR = LLPHONE / TT_HOUSEHOLDS * 1000,
         MPHONE_PR = MPHONE / TT_HOUSEHOLDS * 1000,
         COMPUTER_PR = COMPUTER / TT_HOUSEHOLDS * 1000,
         INTERNET_PR = INTERNET / TT_HOUSEHOLDS * 1000)
```

The summary statistics of the newly derived penetration rates:

```{r}
summary(ict_derived)
```

::: callout-note
Six new fields have been added to the data frame: RADIO_PR, TV_PR, LLPHONE_PR, MPHONE_PR, COMPUTER_PR, and INTERNET_PR.
:::

## 6 Exploratory Data Analysis (EDA)

### 6.1 EDA using statistical graphics

The distribution of the variables (e.g. the number of households with radio) can be plotted with standard EDA graphics.

A histogram is useful to identify the overall distribution of the data values (e.g. left skew, right skew or normal distribution):

```{r}
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

::: callout-note
**Analysis:**

The distribution is heavily right-skewed. Most townships own approximately 1,000 to 5,000 radios, but there is a small handful that owns more than 10,000 radios. 
:::

Boxplot is useful to detect outliers.

```{r}
ggplot(data = ict_derived,
       aes(x = `RADIO`)) +
  geom_boxplot(color = "black",
               fill = "light blue")
```

::: callout-note
**Analysis:**

The boxplot shows there are three outliers. These three townships can also be seen from the histogram, where two are clustered around 14,000 to 17,000, and one isolated near 30,000.
:::

Next, the distribution of the newly derived variables (i.e. radio penetration rate) is plotted:

```{r}
ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

```{r}
#| fig-width: 10
#| fig-height: 5

ggplot(data = ict_derived,
       aes(x = `RADIO_PR`)) +
  geom_boxplot(color = "black",
               fill = "light blue")
```

::: callout-note
**Analysis:**

Compared to the raw `RADIO` count, the distribution of `RADIO_PR` looks much closer to a normal distribution, with most townships sitting between 100 and 300 radios per 1,000 households, and a peak around 225 to 250. There is still a slight right skew, but not as extreme as before.

The boxplot further supports the findings. It is roughly centred, with the median line close to the middle rather than pulled to one edge. Only one outlier remains compared to three before.

This confirms what we expected. Most of the skew and outliers in the raw `RADIO` count were driven by township size instead of actual differences in radio ownership, and standardising to a rate removes most of that effect.
:::

Next, multiple histograms are plotted together to compare the distributions of all six penetration rates in the `ict_derived` data frame.

```{r}
radio <- ggplot(data = ict_derived,
             aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

tv <- ggplot(data = ict_derived,
             aes(x = `TV_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

llphone <- ggplot(data = ict_derived,
             aes(x = `LLPHONE_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

mphone <- ggplot(data = ict_derived,
             aes(x = `MPHONE_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

computer <- ggplot(data = ict_derived,
             aes(x = `COMPUTER_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")

internet <- ggplot(data = ict_derived,
             aes(x = `INTERNET_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue")
```

`ggarrange()` of **ggpubr** groups the six histograms into one figure:

```{r}
ggarrange(radio, tv, llphone, mphone, computer, internet,
          ncol = 3,
          nrow = 2)
```

::: callout-note
**Analysis:**

`RADIO_PR` and `TV_PR` are close to normal, with a central peak and roughly even spread on both sides. On the other hand, the other four (`LLPHONE_PR`, `MPHONE_PR`, `COMPUTER_PR`, and `INTERNET_PR`) are right-skewed. Most townships cluster at the low end, with a long tail of higher values.

This tracks with adoption levels. Radio and TV are widespread enough that most townships land in a similar range. Mobile phone, land-line, computer and internet are rarer and concentrated in fewer townships, hence the skew.
:::

### 6.2 EDA using choropleth map

#### 6.2.1 Joining geospatial data with aspatial data

Before a choropleth map can be prepared, the geospatial object (`shan_sf`) and the aspatial data frame (`ict_derived`) need to be combined into one. This is done with `left_join()` of **dplyr**, with `shan_sf` as the base object and `ict_derived` as the join table. The unique identifier used to join both objects is `TS_PCODE`.

```{r}
shan_sf <- left_join(shan_sf,
                      ict_derived, by = c("TS_PCODE" = "TS_PCODE"))

write_rds(shan_sf, "data/rds/shan_sf.rds")
```

No new output object is created by the join. Instead, the data fields from `ict_derived` are added onto the existing `shan_sf` data frame.

```{r}
shan_sf <- read_rds("data/rds/shan_sf.rds")
```

#### 6.2.2 Preparing a choropleth map

To have a quick look at the distribution of the Radio penetration rate across Shan State at the township level, a choropleth map is prepared using `qtm()` of **tmap**.

```{r}
qtm(shan_sf, "RADIO_PR")
```

To check whether the distribution shown above is biased by the underlying number of households per township, two choropleth maps are prepared side by side: one for the total number of households (`TT_HOUSEHOLDS`) and one for the total number of households with radio (`RADIO`).

```{r}
#| fig-width: 8
#| fig-height: 4

TT_HOUSEHOLDS.map <- tm_shape(shan_sf) +
  tm_fill(col = "TT_HOUSEHOLDS",
          n = 5,
          style = "jenks",
          title = "Total households") +
  tm_borders(alpha = 0.5)

RADIO.map <- tm_shape(shan_sf) +
  tm_fill(col = "RADIO",
          n = 5,
          style = "jenks",
          title = "Number Radio") +
  tm_borders(alpha = 0.5)

tmap_arrange(TT_HOUSEHOLDS.map, RADIO.map,
             asp = NA, ncol = 2)
```

::: callout-note
Notice that townships with relatively more households also show relatively higher radio ownership, which is exactly the bias the penetration rate is meant to remove.
:::

The total households and radio penetration rate are now plotted side by side instead.

```{r}
tm_shape(shan_sf) +
    tm_polygons(c("TT_HOUSEHOLDS", "RADIO_PR"),
                style = "jenks") +
    tm_facets(sync = TRUE, ncol = 2) +
  tm_legend(legend.position = c("right", "bottom")) +
  tm_layout(outer.margins = 0, asp = 0)
```

::: callout-note
**Analysis:**

Once the household-count bias is removed, the two maps no longer agree. Several townships with a moderate number of households show among the highest radio penetration rates, while some of the larger townships by household count show relatively low penetration rates.
:::

## 6 Correlation Analysis

Before performing cluster analysis, it is important to ensure that the clustering variables are not highly correlated. 

In this section, `corrplot.mixed()` of [**corrplot**](https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html) is used to visualise and analyse the correlation of the input variables.

```{r}
cluster_vars.cor = cor(ict_derived[,12:17])
corrplot.mixed(cluster_vars.cor,
         lower = "ellipse",
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

::: callout-important
The correlation plot above shows that `COMPUTER_PR` and `INTERNET_PR` are highly correlated (r = 0.87). This suggest that only one of them should be used in the cluster analysis instead of both.
:::

## 7 Hierarchy Cluster Analysis

### 7.1 Extracting clustering variables

The clustering variables are extracted from the `shan_sf` simple feature object into a data frame.

```{r}
cluster_vars <- shan_sf %>%
  st_set_geometry(NULL) %>%
  select("TS.x", "RADIO_PR", "TV_PR", "LLPHONE_PR", "MPHONE_PR", "COMPUTER_PR")
head(cluster_vars, 10)
```

::: callout-note
Notice that the final clustering variables list does not include `INTERNET_PR` because it is highly correlated with `COMPUTER_PR`.
:::

Next, the rows are changed to be labelled by township name instead of row number:

```{r}
row.names(cluster_vars) <- cluster_vars$"TS.x"
head(cluster_vars, 10)
```

The `TS.x` field is then dropped, since the township name is now the row name and no longer needed as a column:

```{r}
shan_ict <- select(cluster_vars, c(2:6))
head(shan_ict, 10)
```

```{r}
write_rds(shan_ict, "data/rds/shan_ict.rds")
shan_ict <- read_rds("data/rds/shan_ict.rds")
```

### 7.2 Data Standardisation

In general, multiple variables are used in cluster analysis. It is not unusual for their value ranges to differ. As seen earlier, `TV_PR` ranges into the hundreds while `COMPUTER_PR` stays under 100. To avoid the cluster analysis from being biased toward clustering variables with large values, the input variables are standardised before clustering.

### 7.3 Min-Max standardisation

In the code chunk below, `normalize()` of [**heatmaply**](https://cran.r-project.org/web/packages/heatmaply/index.html) is used to standardise the clustering variables with the Min-Max method. The `summary()` then shows the summary statistics of the standardised clustering variables.

```{r}
shan_ict.std <- normalize(shan_ict)
summary(shan_ict.std)
```

The min-max standardised clustering variables now **range from 0 to 1**.

### 7.4 Z-score standardisation

Z-score standardisation can be performed with [scale()](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) of base R. The code below standardises the clustering variables using the Z-score method.

```{r}
shan_ict.z <- scale(shan_ict)
describe(shan_ict.z)
```

::: callout-note
Notice the mean and standard deviation of the Z-score standardised clustering variables are 0 and 1 respectively.
:::

::: callout-warning
The Z-score standardisation method should only be used if the variables can be assumed to come from some normal distribution.
:::

### 7.5 Visualising the standardised clustering variables

Besides reviewing the summary statistics, it is also a good practice to visualise the distribution of the standardised clustering variables graphically.

The code chunk below plots the scaled `RADIO_PR` field:

```{r}
#| fig-width: 12
#| fig-height: 4

r <- ggplot(data = ict_derived,
             aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data = shan_ict_s_df,
       aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data = shan_ict_z_df,
       aes(x = `RADIO_PR`)) +
  geom_histogram(bins = 20,
                 color = "black",
                 fill = "light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

::: callout-note
**Analysis:**

From the above histograms, it can be observed that standardisation changes scale but not the shape. The three histograms look identical aside from their axis values. Since min-max and Z-score standardisation are both linear transformations, they rescale the data but do not alter its distribution. Whatever skew is present in the raw `RADIO_PR` values remains after standardisation, just measured on a different scale.
:::

The same comparison is repeated below using density plots instead of histograms:

```{r}
#| fig-width: 12
#| fig-height: 4

r <- ggplot(data = ict_derived,
             aes(x = `RADIO_PR`)) +
  geom_density(color = "black",
               fill = "light blue") +
  ggtitle("Raw values without standardisation")

shan_ict_s_df <- as.data.frame(shan_ict.std)
s <- ggplot(data = shan_ict_s_df,
       aes(x = `RADIO_PR`)) +
  geom_density(color = "black",
               fill = "light blue") +
  ggtitle("Min-Max Standardisation")

shan_ict_z_df <- as.data.frame(shan_ict.z)
z <- ggplot(data = shan_ict_z_df,
       aes(x = `RADIO_PR`)) +
  geom_density(color = "black",
               fill = "light blue") +
  ggtitle("Z-score Standardisation")

ggarrange(r, s, z,
          ncol = 3,
          nrow = 1)
```

### 7.6 Computing proximity matrix

Many R packages provide functions to calculate distance matrix. One of the ways to compute proximity matrix is by using [dist()](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/dist.html) of base R.

`dist()` supports six distance methods: euclidean, maximum, manhattan, canberra, binary and minkowski. The default method is **euclidean**.

```{r}
proxmat <- dist(shan_ict, method = 'euclidean')
```

The content of `proxmat` can be listed for inspection:

```{r}
#| results: hide
proxmat
```

### 7.7 Computing hierarchical clustering

Several R packages provide hierarchical clustering functions. In this exercise, [hclust()](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/hclust.html) of R stats is used.

`hclust()` uses an agglomeration method to compute the cluster. Eight clustering algorithms are supported: ward.D, ward.D2, single, complete, average (UPGMA), mcquitty (WPGMA), median (WPGMC) and centroid (UPGMC).

The code chunk below performs hierarchical cluster analysis using the **ward.D method**. The output is stored in an object of class `hclust`, which describes the tree produced by the clustering process.

```{r}
hclust_ward <- hclust(proxmat, method = 'ward.D')
```

The tree can be plotted with `plot()` of R Graphics:

```{r}
#| fig-width: 10
#| fig-height: 6

plot(hclust_ward, cex = 0.6)
```

### 7.8 Selecting the optimal clustering algorithm

One of the challenges in hierarchical clustering is identifying which method produces the strongest clustering structure. This can be solved by using `agnes()` of [**cluster**](https://cran.r-project.org/web/packages/cluster/) package. It functions like `hclust()`, but also returns the agglomerative coefficient, which measures the amount of clustering structure found (values closer to 1 suggest strong clustering structure).

The code chunk below computes the agglomerative coefficients of four hierarchical clustering algorithms.

```{r}
m <- c("average", "single", "complete", "ward")
names(m) <- c("average", "single", "complete", "ward")

ac <- function(x) {
  agnes(shan_ict, method = x)$ac
}

map_dbl(m, ac)
```

With reference to the output, **Ward's method** provides the strongest clustering structure among the four methods assessed. Ward's method is therefore used in the subsequent analysis.

### 7.9 Determining Optimal Clusters

Another challenge in cluster analysis is determining the optimal number of clusters to retain. 

There are three commonly used methods:

- [Elbow method](https://en.wikipedia.org/wiki/Elbow_method_(clustering))
- [Average silhouette method](https://www.sciencedirect.com/science/article/pii/0377042787901257?via%3Dihub)
- [Gap statistic method](https://hastie.su.domains/Papers/gap.pdf)

#### 7.9.1 Gap statistic method

The gap statistic compares the total intra-cluster variation for different values of k against their expected values under a null reference distribution of the data. The estimated optimal number of clusters is the value that maximises the gap statistic (i.e. yields the largest gap statistic). This indicates that the clustering structure is furthest from a random uniform distribution of points.

`clusGap()` of **cluster** is used to compute the gap statistic.

```{r}
set.seed(12345)
gap_stat <- clusGap(shan_ict,
                    FUN = hcut,
                    nstart = 25,
                    K.max = 10,
                    B = 50)
print(gap_stat, method = "firstmax")
```

::: callout-note
The `hcut()` function used above is from *[*factoextra**](https://rpkgs.datanovia.com/factoextra/index.html) package.
:::

The result can be visualised using `fviz_gap_stat()` of **factoextra**:

```{r}
fviz_gap_stat(gap_stat)
```

::: callout-note
**Analysis:**

With reference to the gap statistic graph above, the recommended number of cluster to retain is 1. However, it is not logical to retain only one cluster. Looking at the graph more closely, the 6-cluster solution gives the largest gap statistic among the remaining options and is the next best cluster to pick.
:::

::: callout-tip
In addition to these commonly used approaches, the [**NbClust**](https://cran.r-project.org/web/packages/NbClust/) package, published by Charrad et al. (2014), provides 30 indices for determining the number of clusters and proposes the best clustering scheme across combinations of cluster counts, distance measures and clustering methods.
:::

### 7.10 Interpreting the dendrograms

In the dendrogram, each leaf corresponds to one observation (township). Moving up the tree, observations that are similar to each other are combined into branches, which are themselves fused at a higher height.

The height of a fusion (i.e. vertical axis) indicates the dissimilarity between two branches. The higher the height of the fusion, the less similar the observations are. Note that conclusions about the proximity of two observations can only be drawn from the height at which their branches first fuse, not from their horizontal proximity in the plot.

A border can be drawn around the selected clusters using `rect.hclust()` of R stats. The argument `border` is used to specify the rectangle colours.

```{r}
#| fig-width: 10
#| fig-height: 6

plot(hclust_ward, cex = 0.6)
rect.hclust(hclust_ward,
            k = 6,
            border = 2:5)
```

### 7.11 Visually-driven hierarchical clustering analysis

This section uses [**heatmaply**](https://cran.r-project.org/web/packages/heatmaply/index.html) to perform a visually-driven hierarchical clustering analysis, building both an interactive and a static cluster heatmap.

#### 7.11.1 Transforming the data frame into a matrix

The clustering data is currently a data frame, but `heatmaply()` requires a data matrix.

The code chunk below will be used to transform `shan_ict` data frame into a data matrix.

```{r}
shan_ict_mat <- data.matrix(shan_ict)
```

#### Plotting interactive cluster heatmap using heatmaply()

`heatmaply()` of **heatmaply** package is used to build an interactive cluster heatmap.

```{r}
#| fig-width: 10
#| fig-height: 10

heatmaply(normalize(shan_ict_mat),
          Colv = NA,
          dist_method = "euclidean",
          hclust_method = "ward.D",
          seriate = "OLO",
          colors = Blues,
          k_row = 6,
          margins = c(NA, 200, 60, NA),
          fontsize_row = 4,
          fontsize_col = 5,
          main = "Geographic Segmentation of Shan State by ICT indicators",
          xlab = "ICT Indicators",
          ylab = "Townships of Shan State"
          )
```

### 7.12 Mapping the clusters formed

With close examination of the dendrogram above, six clusters are retained.

[cutree()](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/cutree.html) of R base is used to derive a 6-cluster model:

```{r}
groups <- as.factor(cutree(hclust_ward, k = 6))
```

The output (`groups`) is a list object. 

To visualise the clusters, it needs to be appended onto the `shan_sf` simple feature object. This join can be done in three steps:

- `groups` is converted into a matrix;
- `cbind()` appends the `groups` matrix onto `shan_sf` to produce a new simple feature object called `shan_sf_cluster`; and
- `rename()` of **dplyr** is used to rename `as.matrix.groups.` field as `CLUSTER`.

```{r}
shan_sf_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

Next, `qtm()` of **tmap** plots the choropleth map showing the clusters formed.

```{r}
qtm(shan_sf_cluster, "CLUSTER")
```

::: callout-note
**Analysis:**

The choropleth map reveals that the clusters are very fragmented. This is one of the major limitations of using a non-spatial clustering algorithm, such as hierarchical cluster analysis, on geographically referenced data. The attribute similarity it optimises for has no regard for whether the resulting groups are geographically contiguous.
:::

## 8 Spatially Constrained Clustering: SKATER Approach

In this section, a spatially constrained cluster is derived using [skater()](https://r-spatial.github.io/spdep/reference/skater.html) of [**spdep**](https://r-spatial.github.io/spdep/index.html) package.

### 8.1 Converting into SpatialPolygonsDataFrame

`shan_sf` is first converted into a SpatialPolygonsDataFrame, since the SKATER function only supports **sp** objects such as SpatialPolygonsDataFrame.

`as_Spatial()` of **sf** is used to convert `shan_sf` into a SpatialPolygonsDataFrame called `shan_sp`.

```{r}
shan_sp <- as_Spatial(shan_sf)
```

### 8.2 Computing Neighbour List

Next, `poly2nb()` of **spdep** is used to compute the neighbours list from the polygon list.

```{r}
shan.nb <- poly2nb(shan_sp)
summary(shan.nb)
```

The neighbours list can be plotted on top of `shan_sp`. The boundaries are plotted first, followed by the neighbour list object, with coordinates applied to the original SpatialPolygonsDataFrame to extract the centroids of the polygons as nodes for the graph. The network is plotted in blue with `add = TRUE` on top of the boundaries.

```{r}
coords <- st_coordinates(
  st_centroid(st_geometry(shan_sf)))
```

```{r}
#| fig-width: 8
#| fig-height: 6

plot(st_geometry(shan_sf),
     border = grey(.5))
plot(shan.nb,
     coords,
     col = "blue",
     add = TRUE)
```

::: callout.note
The boundary map is plotted first and the network second, since the plotting area is determined by the first plot drawn. If the network was plotted before the boundaries, and the boundary map extends further than the graph, some areas would be clipped.
:::

### 8.3 Computing edge costs

`nbcosts()` of **spdep** is used to compute the cost of each edge. It is the distance between its nodes. This function computes that distance using a data frame with an observation vector at each node.

```{r}
lcosts <- nbcosts(shan.nb, shan_ict)
```

For each observation, this gives the pairwise dissimilarity between its values on the five clustering variables and the values of its neighbouring observation (from the neighbour list). Basically, this is the notion of a generalised weight for a spatial weights matrix.

These costs are then incorporated into a weights object in the same way as computing inverse-distance weights. The neighbour list is converted into a list weights object by specifying the newly computed `lcosts` as the weights. 

In order to achieve this, `nb2listw()` of **spdep** is used, with `style` set to **"B"** to ensure the cost values are not row-standardised.

```{r}
shan.w <- nb2listw(shan.nb,
                   lcosts,
                   style = "B")
summary(shan.w)
```

### 8.4 Computing minimum spanning tree

The minimum spanning tree is computed with [mstree()](https://r-spatial.github.io/spdep/reference/mstree.html) of **spdep**:

```{r}
shan.mst <- mstree(shan.w)
```

Its class and dimensions can be checked:

```{r}
class(shan.mst)
```

```{r}
dim(shan.mst)
```

::: callout-note
The dimension is 54, and not 55. This is because the minimum spanning tree has n-1 edges (links) in order to traverse all n nodes.
:::

The content of `shan.mst` can be inspected using `head()`:

```{r}
head(shan.mst)
```

The MST is plotted together with the township boundaries, including the observation numbers of the nodes:

```{r}
#| fig-width: 8
#| fig-height: 6

plot(st_geometry(shan_sf),
                 border = gray(.5))
plot.mst(shan.mst,
         coords,
         col = "blue",
         cex.lab = 0.7,
         cex.circles = 0.005,
         add = TRUE)
```

This shows how the initial neighbour list is simplified down to a single edge connecting each node, while still passing through every node in the graph.

### 8.5 Computing spatially constrained clusters using SKATER method

The code chunk below compute the spatially constrained cluster using [skater()](https://r-spatial.github.io/spdep/reference/skater.html) of **spdep** package.

```{r}
clust6 <- spdep::skater(edges = shan.mst[,1:2],
                 data = shan_ict,
                 method = "euclidean",
                 ncuts = 5)
```

The `skater()` takes three mandatory arguments: 

- the first two columns of the MST matrix (not the cost column)
- the data matrix (to update the costs as units are grouped) 
- the number of cuts

The number of cuts is set to **one less than the number of clusters**, so the value specified is **not** the number of clusters, but the number of cuts made in the graph.

The result is an object of class **skater**. Its contents can be examined using `str()`.

```{r}
str(clust6)
```

The most interesting component of this structure is the groups vector containing the cluster label each observation belongs to (the labels themselves are arbitrary). This is followed by a detailed summary for each cluster in the `edges.groups` list. Sum-of-squares measures are given as `ssto` for the total and `ssw` for the effect of each cut on the overall criterion.

The cluster assignment can be checked:

```{r}
ccs6 <- clust6$groups
ccs6
```

The number of observations in each cluster can be found with `table()`.

```{r}
table(ccs6)
```

::: callout-note
The cluster sizes are highly uneven. Clusters 1, 2 and 3 account for 51 of the 55 townships (22, 18 and 11 respectively), while clusters 4, 5 and 6 are very small, with 2, 1 and 1 townships respectively.
:::

Finally, the pruned tree showing the clusters can be plotted on top of the township boundaries.

```{r}
#| fig-width: 8
#| fig-height: 6

plot(st_geometry(shan_sf),
     border = gray(.5))
plot(clust6,
     coords,
     cex.lab = .7,
     groups.colors = c("red", "green", "blue", "brown", "pink", "purple"),
     cex.circles = 0.005,
     add = TRUE)
```

### 8.6 Visualising the clusters in choropleth map

The code chunk below is used to plot the newly derived clusters by using SKATER method.

```{r}
#| fig-width: 8
#| fig-height: 6

groups_mat <- as.matrix(clust6$groups)
shan_sf_spatialcluster <- cbind(shan_sf_cluster, as.factor(groups_mat)) %>%
  rename(`SP_CLUSTER` = `as.factor.groups_mat.`)
qtm(shan_sf_spatialcluster, "SP_CLUSTER")
```

For easier comparison, the hierarchical clustering and spatially constrained clustering maps are placed side by side:

```{r}
hclust.map <- qtm(shan_sf_cluster,
                  "CLUSTER") +
  tm_borders(alpha = 0.5)

shclust.map <- qtm(shan_sf_spatialcluster,
                   "SP_CLUSTER") +
  tm_borders(alpha = 0.5)

tmap_arrange(hclust.map, shclust.map,
             asp = NA, ncol = 2)
```

::: callout-note
**Analysis:**

Compared to the fragmented pattern from hierarchical clustering, the SKATER clusters form contiguous regions on the map, since the spatial constraint forces each cluster to consist of townships that are directly connected through the minimum spanning tree.
:::

## 9 Spatially Constrained Clustering: ClustGeo Method

This section uses functions provided by **ClustGeo** to perform both non-spatially constrained and spatially constrained hierarchical cluster analysis.

### 9.1 Introduction to ClustGeo package

**ClustGeo** is an R package designed to support spatially constrained cluster analysis. It provides a Ward-like hierarchical clustering algorithm called `hclustgeo()` that incorporates spatial or geographical constraints.

The algorithm uses two dissimilarity matrices, D0 and D1, along with a mixing parameter alpha, where the value of alpha must be a real number between 0 and 1. D0 can be non-Euclidean and the weights of the observations can be non-uniform. It gives the dissimilarities in the **attribute/clustering variable space**. D1, on the other hand, gives the dissimilarities in the **constraint space**. The criterion minimised at each stage is a convex combination of the homogeneity criterion calculated with D0 and the homogeneity criterion calculated with D1.

The idea is to find a value of alpha that increases spatial contiguity without deteriorating the quality of the solution based on the variables of interest. This is supported by a function called `choicealpha()`.

### 9.2 Ward-like hierarchical clustering: ClustGeo

**ClustGeo** provides `hclustgeo()` to perform Ward-like hierarchical clustering, similar to `hclust()` used in the earlier section.

To perform non-spatially constrained hierarchical clustering, only a dissimilarity matrix needs to be supplied.

```{r}
#| fig-width: 8
#| fig-height: 6

nongeo_cluster <- hclustgeo(proxmat)
plot(nongeo_cluster, cex = 0.5)
rect.hclust(nongeo_cluster,
            k = 6,
            border = 2:5)
```

::: callout-note
The dissimilarity matrix must be an object of class `dist`, i.e. an object obtained with the function `dist()`.
:::

#### 9.2.1 Mapping the clusters formed

The clusters can be plotted on a categorical choropleth map using the same steps as in the earlier hierarchical clustering section (Section 7.12).

```{r}
groups <- as.factor(cutree(nongeo_cluster, k = 6))
```

```{r}
shan_sf_ngeo_cluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

```{r}
qtm(shan_sf_ngeo_cluster, "CLUSTER")
```

### 9.3 Spatially Constrained Hierarchical Clustering

Before spatially constrained hierarchical clustering can be performed, a spatial distance matrix is derived with `st_distance()` of **sf** package.

```{r}
dist <- st_distance(shan_sf, shan_sf)
distmat <- as.dist(dist)
```

::: callout-note
`as.dist()` converts the data frame into a matrix.
:::

Next, `choicealpha()` is used to determine a suitable value for the mixing parameter alpha.

```{r}
#| fig-width: 8
#| fig-height: 6

cr <- choicealpha(proxmat, distmat, range.alpha = seq(0, 1, 0.1), K = 6, graph = TRUE)
```

With reference to the graphs above, alpha = 0.2 is used.

```{r}
clustG <- hclustgeo(proxmat, distmat, alpha = 0.2)
```

Next, `cutree()` is used to derive the cluster object.

```{r}
groups <- as.factor(cutree(clustG, k = 6))
```

The group list is joined back with the `shan_sf` polygon feature data frame.

```{r}
shan_sf_Gcluster <- cbind(shan_sf, as.matrix(groups)) %>%
  rename(`CLUSTER` = `as.matrix.groups.`)
```

The map of the newly delineated spatially constrained clusters:

```{r}
qtm(shan_sf_Gcluster, "CLUSTER")
```

## 10 Visual Interpretation of Clusters

### 10.1 Visualising individual clustering variable

The code chunk below reveals the distribution of a clustering variable (`RADIO_PR`) by cluster.

```{r}
ggplot(data = shan_sf_ngeo_cluster,
       aes(x = CLUSTER, y = RADIO_PR)) +
  geom_boxplot()
```

::: callout-note
**Analysis:**

The boxplot reveals Cluster 3 displays the highest median radio penetration rate. This is followed by Cluster 2, 1, and 4. Clusters 5 and 6 are the lowest, both with medians around 90 to 95.

The clusters also differ in spread. Cluster 4 has the widest range (roughly 55 to 360), suggesting its member townships are quite varied in radio ownership despite sharing a cluster. Clusters 1 and 2 are comparatively tight. Two outliers stand out: one in Cluster 1 near 420 and one in Cluster 3 near 480, both single townships well above the rest of their cluster.
:::

### 10.2 Multivariate Visualisation

Parallel coordinate plots are an effective way to reveal clustering variables by cluster. `ggparcoord()` of **GGally** package is used below.

```{r}
#| fig-width: 14
#| fig-height: 6

ggparcoord(data = shan_sf_ngeo_cluster, 
           columns = c(17:21), 
           scale = "globalminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of ICT Variables by Cluster") +
  facet_grid(~ CLUSTER) + 
  theme(axis.text.x = element_text(angle = 30))
```

::: callout-note
**Analysis:**

The parallel coordinate plot above reveals that households in Cluster 4 townships tend to own the highest number of TV and mobile-phone. On the other hand, households in Cluster 5 tends to own the lowest of all the five ICT.
:::

::: callout-tip
The `scale` argument of `ggparcoord()` provides several methods to scale the clustering variables. They are:

- **std**: univariately, subtract the mean and divide by the standard deviation.
- **robust**: univariately, subtract the median and divide by the median absolute deviation.
- **uniminmax**: univariately, scale so the minimum of the variable is zero and the maximum is one.
- **globalminmax**: no scaling is done; the range of the graphs is defined by the global minimum and global maximum.
- **center**: use `uniminmax` to standardise vertical height, then centre each variable at a value specified by the `scaleSummary` parameter.
- **centerObs**: use `uniminmax` to standardise vertical height, then centre each variable at the value of the observation specified by the `centerObsID` parameter.

There is no single best scaling method. It is worth exploring them and choosing the one that best suits the analysis.
:::

Summary statistics can also be computed to complement the visual interpretation. 

In the code chunk below, `group_by()` and `summarise()` of **dplyr** are used to derive mean values of the clustering variables.

```{r}
shan_sf_ngeo_cluster %>%
  st_set_geometry(NULL) %>%
  group_by(CLUSTER) %>%
  summarise(mean_RADIO_PR = mean(RADIO_PR),
            mean_TV_PR = mean(TV_PR),
            mean_LLPHONE_PR = mean(LLPHONE_PR),
            mean_MPHONE_PR = mean(MPHONE_PR),
            mean_COMPUTER_PR = mean(COMPUTER_PR))
```

