---
title: "Hands-on Exercise 2B: Second-order Spatial Point Pattern Analysis (2nd-SPPA)"
---

## 1 Overview

**Second-order Spatial Point Pattern Analysis (2nd-SPPA)** examines the spatial relationships between points in a pattern, specifically focusing on how the presence of one point influences the location of others. It goes beyond simply describing the overall density of points (first-order effects) by investigating clustering, dispersion, or randomness at various spatial scales.

This exercise uses [**spatstat**](https://cran.r-project.org/web/packages/spatstat/) to discover the spatial point processes of childcare centres in Singapore, and will focus on two specific questions:

- Are the childcare centres in Singapore randomly distributed throughout the country?
- If the answer is not, then the next logical question is: where are the locations with higher concentration of childcare centres?

## 2 The Data

Two data sets are used for this exercise:

| Dataset | Format | Source | Description |
|---|---|---|---|
| Child Care Services | GeoJSON | [data.gov.sg](https://data.gov.sg/) | Point feature data providing the location and attribute information of childcare centres in Singapore. |
| Master Plan 2019 Subzone Boundary (No Sea) | GeoJSON | [data.gov.sg](https://data.gov.sg/) | Polygon feature data providing the URA 2019 Master Plan Planning Subzone boundaries. |

Both data sets are the same as those used in [**Hands-on Exercise 2A**](https://isss-626-gaa-one.vercel.app/Hands-on_Ex/Hands-on_Ex02A/Hands-on_Ex02A.html), which covers 1st-order SPPA.

## 3 Installing and Loading the R packages

Four R packages will be used for this exercise:

| Package | Purpose | Use Case in Exercise |
|---|---|---|
| [**sf**](https://r-spatial.github.io/sf/) | Imports, manages, and processes vector-based geospatial data. | Handling vector geospatial data in R. |
| [**spatstat**](https://spatstat.org/) | Provides tools for point pattern analysis. | Performing 1st-order spatial point pattern analysis and deriving kernel density estimation (KDE). |
| [**tmap**](https://cran.r-project.org/web/packages/tmap/index.html) | Creates static and interactive thematic maps using cartographic-quality elements. | Plotting the KDE surface as a static map, and the childcare point locations as an interactive map via [leaflet](https://leafletjs.com/). |
| [**tidyverse**](https://www.tidyverse.org/) | Family of R packages for modern data science (data manipulation, transformation). | Data wrangling and manipulation with `mutate()`, `filter()`, and pipe operators throughout the exercise. |

The code chunk below installs and loads the four R packages:

```{r}
pacman::p_load(sf, spatstat, tmap, tidyverse)
```

## 4 Data Import and Preparation

Data import and preparation for this exercise reuse the objects prepared in [Hands-on Exercise 2A](https://isss-626-gaa-one.vercel.app/Hands-on_Ex/Hands-on_Ex02A/Hands-on_Ex02A.html): the cleaned boundary data (`mpsz_cl`) and the jittered childcare point pattern (`childcare_ppp_jit`). Both were saved there as `.rds` files and copied into this exercise's own `data` folder. Full details of the import, CRS transformation, and jittering steps are available under the following Hands-on Exercise 2A sections:

- [Importing and Wrangling Geospatial Data Sets](https://isss-626-gaa-one.vercel.app/Hands-on_Ex/Hands-on_Ex02A/Hands-on_Ex02A.html#importing-and-wrangling-geospatial-data-sets)
- [Geospatial Data Wrangling](https://isss-626-gaa-one.vercel.app/Hands-on_Ex/Hands-on_Ex02A/Hands-on_Ex02A.html#geospatial-data-wrangling)

The code chunk below loads both objects, then extracts the same four planning areas used in Hands-on Exercise 2A — Punggol, Tampines, Choa Chu Kang, and Jurong West:

```{r}
mpsz_cl <- read_rds("data/mpsz_cl.rds")
childcare_ppp_jit <- read_rds("data/childcare_ppp_jit.rds")

pg <- mpsz_cl %>% filter(PLN_AREA_N == "PUNGGOL")
tm <- mpsz_cl %>% filter(PLN_AREA_N == "TAMPINES")
ck <- mpsz_cl %>% filter(PLN_AREA_N == "CHOA CHU KANG")
jw <- mpsz_cl %>% filter(PLN_AREA_N == "JURONG WEST")

pg_owin <- as.owin(pg)
tm_owin <- as.owin(tm)
ck_owin <- as.owin(ck)
jw_owin <- as.owin(jw)

childcare_pg_ppp <- childcare_ppp_jit[pg_owin]
childcare_tm_ppp <- childcare_ppp_jit[tm_owin]
childcare_ck_ppp <- childcare_ppp_jit[ck_owin]
childcare_jw_ppp <- childcare_ppp_jit[jw_owin]
```

The four planning areas are plotted below to confirm the extraction:

```{r}
#| fig-width: 10
#| fig-height: 10
par(mfrow = c(2, 2))
plot(unmark(childcare_pg_ppp), main = "Punggol")
plot(unmark(childcare_tm_ppp), main = "Tampines")
plot(unmark(childcare_ck_ppp), main = "Choa Chu Kang")
plot(unmark(childcare_jw_ppp), main = "Jurong West")
```



