---
title: "In-class Exercise 5"
date: "September 26, 2026"
date-modified: today
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1 Installing and Loading R Packages

Four packages are used in this exercise:

- **sf** imports and handles the geospatial data
- **sfdep** computes the spatial weights and the local spatial autocorrelation statistics
- **tmap** produces the choropleth maps
- **tidyverse** supports attribute data handling

```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

## 2 Reproducibility

```{r}
set.seed(123)
```

## 3 Getting the Data into R

### 3.1 Importing the Shapefile

The shapefile is imported with `st_read()` of **sf**, which returns a simple features object.

```{r}
hunan <- st_read(dsn = "data/geospatial",
                 layer = "Hunan")
```

### 3.2 Importing the CSV File

`Hunan_2012.csv` is imported with `read_csv()` of **readr**, which returns a data frame class.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
```

### 3.3 Performing relational join

The attribute table of `hunan` is extended with the fields of `hunan2012` using `left_join()` of **dplyr**.

```{r}
hunan_GDPPC <- left_join(hunan, hunan2012) %>%
  select(1:4, 7, 15, 16)
```

## 4 Deriving Queen's Continguity Weights: sfdep methods

The code chunk below is used to compute the Queen's contiguity weights.

```{r}
wm_q <- hunan_GDPPC %>%
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb,
                         style = "W"),
         .before = 1)
```

## 5 Computing Global Moran's I

In the code chunk below, `global_moran()` function is used to compute the Moran's I value. Different from spdep package, the output is a tibble data frame.

```{r}
moranI <- global_moran(wm_q$GDPPC,
                       wm_q$nb,
                       wm_q$wt)
glimpse(moranI)
```

## 6 Performing Global Moran's I Permutation Test

```{r}
global_moran_perm(wm_q$GDPPC,
                  wm_q$nb,
                  wm_q$wt,
                  nsim = 99)
```

::: callout-tip
## Reminder

- It is always a good practice to use `set.seed()` before performing simulation. This is to ensure that the computation is reproducible.
- The numbers of simulation is always equal to nsim + 1.
:::

## 7 Computing local Moran's I

```{r}
lisa <- wm_q %>%
  mutate(local_moran = local_moran(
    GDPPC, nb, wt, nsim = 99), 
    .before = 1) %>%
  unnest(local_moran)
```

## 8 Plotting LISA Map

```{r}
lisa_sig <- lisa %>%
  filter(p_ii_sim < 0.05)
tmap_mode("plot")
tm_shape(lisa) +
  tm_polygons() +
  tm_borders(fill_alpha = 0.5) +
  tm_shape(lisa_sig) +
  tm_polygons(fill = "mean",
              fill.legend = tm_legend(
                text.size = 1.15,
                width = 6)) +
  tm_borders(fill_alpha = 0.4) +
  tm_layout(legend.title.size = 1.25)
```
