---
title: "Hands-on Exercise 1B: Thematic Mapping and GeoVisualisation with R"
---

## 1 Overview

Thematic mapping involves the use of map symbols to visualize selected properties of geographic features that are not naturally visible, such as population, temperature, crime rate, and property prices.

Geovisualisation, on the other hand, is the process of using visual representations and cartographic techniques to explore, analyze, and communicate geospatial data. It combines elements of cartography, computer science, and information visualization to enhance spatial understanding and knowledge discovery.

This exercise explores how to create functional and truthful choropleth maps using the **tmap** package.

## 2 Getting Started

The key R package used for this exercise is [**tmap**](https://cran.r-project.org/web/packages/tmap/).

Beside **tmap**, five other R packages are used:

- [**readr**](https://readr.tidyverse.org/) for importing delimited text file,
- [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,
- [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data and
- [**sf**](https://cran.r-project.org/web/packages/sf/index.html) for handling geospatial data.

::: callout-note
Among these, **readr**, **tidyr**, and **dplyr** are part of the **tidyverse package.**

Hence, only **tidyverse** needs to be installed, instead of **readr**, **tidyr**, and **dplyr** individually.
:::

The code chunk below installs and loads these packages in RStudio.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## 3 Importing Data into R

### 3.1 The Data

Two datasets are used to create the choropleth map:

| Dataset | Format | Source | Description |
|------------------|------------------|------------------|------------------|
| Master Plan 2019 Subzone Boundary (No Sea) | GeoJSON | [data.gov.sg](https://data.gov.sg) | Geospatial data consisting of Singapore's geographical boundary at the planning subzone level, based on URA Master Plan 2019 |
| Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2024 (`respopagesextod2024.csv`) | CSV | [Department of Statistics, Singapore](https://www.singstat.gov.sg) | Aspatial data. Contains no coordinate values, but the `PA` and `SZ` fields serve as unique identifiers for georeferencing to the Master Plan 2019 Subzone Boundary |

### 3.2 Importing Geospatial Data into R

The code chunk below uses the `st_read()` function of **sf** package to import Master Plan 2019 Subzone Boundary (No Sea) into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read("data/geospatial/MasterPlan2019SubzoneBoundaryNoSeaGEOJSON.geojson") %>%
  st_transform(crs = 3414)
```

The imported `mpsz` contains 13 fields. This exercise will focus on four attribute fields: `REGION_N`, `PLN_AREA_N`, `SUBZONE_N`, and `SUBZONE_C`, along with the geometry column.

```{r}
mpsz <- mpsz %>%
  select(REGION_N, PLN_AREA_N, SUBZONE_N, SUBZONE_C, geometry)
```

```{r}
mpsz
```

::: callout-note
**Why only the first 10 records are displayed?**

**sf** objects print only the first 10 records by default to avoid flooding the console with all records.
:::

### 3.3 Importing Attribute Data into R

The `respopagesextod2024.csv` file is imported into a tibble dataframe called `popdata`, using `read_csv()` of the **readr** package.

```{r}
#| message: false
popdata <- read_csv("data/aspatial/respopagesextod2024.csv")
```

### 3.4 Data Preparation

Before a thematic map can be prepared, a data table with year 2024 values is required. The data table includes the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, and DEPENDENCY.

- **YOUNG:** age group 0 to 4 until age group 20 to 24
- **ECONOMY ACTIVE:** age group 25-29 until age group 60-64
- **AGED:** age group 65 and above
- **TOTAL:** all age groups
- **DEPENDENCY:** the ratio between young and aged against the economy active group

#### 3.4.1 Data Wrangling

The following data wrangling and transformation functions are used:

- *pivot_wider()* of the **tidyr** package, and
- *mutate()*, *filter()*, *group_by()*, and *select()* of the **dplyr** package

```{r}
#| message: false

# Step 1: Group the data by PA (Planning Area), SZ (Subzone), and AG (Age Group)
popdata2024 <- popdata %>%
  group_by(PA, SZ, AG)

# Step 2: Summarise the data by calculating the total population (POP) for each group
popdata2024 <- popdata2024 %>%
  summarise(POP = sum(Pop)) %>%
  ungroup()

# Step 3: Reshape the data from long to wide format, with age groups as column names
popdata2024 <- popdata2024 %>%
  pivot_wider(names_from = AG, values_from = POP)

# Step 4: Calculate the YOUNG population (sum of selected age groups)
popdata2024 <- popdata2024 %>%
  mutate(YOUNG = rowSums(.[3:6]) + rowSums(.[14]))

# Step 5: Calculate the ECONOMY ACTIVE population (sum of selected age groups)
popdata2024 <- popdata2024 %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:13]) + rowSums(.[15]))

# Step 6: Calculate the AGED population (sum of selected age groups)
popdata2024 <- popdata2024 %>%
  mutate(AGED = rowSums(.[16:21]))

# Step 7: Calculate the TOTAL population (sum of all age groups)
popdata2024 <- popdata2024 %>%
  mutate(TOTAL = rowSums(.[3:21]))

# Step 8: Calculate the DEPENDENCY ratio (ratio of YOUNG and AGED to ECONOMY ACTIVE)
popdata2024 <- popdata2024 %>%
  mutate(DEPENDENCY = (YOUNG + AGED) / `ECONOMY ACTIVE`)

# Step 9: Select the relevant columns to be retained in the final dataset
popdata2024 <- popdata2024 %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)
```

#### 3.4.2 Joining the attribute data and geospatial data

Before the georelational join, an extra step is required to convert the values in the PA and SZ fields to uppercase. This is necessary because the PA and SZ fields contain a mix of upper- and lowercase, while SUBZONE_N and PLN_AREA_N are entirely uppercase.

```{r}
popdata2024 <- popdata2024 %>%
  mutate_at(.vars = vars(PA, SZ),
            .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` of **dplyr** is used to join the geographical data and attribute table using planning subzone name, e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2024 <- left_join(mpsz, popdata2024,
                           by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
`left_join()` of **dplyr** is used with `mpsz` as the left data table, to ensure that the output will be a simple features data frame.
:::

```{r}
write_rds(mpsz_pop2024, "data/rds/mpsz_pop2024.rds")
```

## 4 Choropleth Mapping Geospatial Data Using tmap
