---
title: "Take-home Exercise 1: Geospatial Analytics for Public Safety"
subtitle: "Spatial and Spatio-Temporal Point Pattern Analysis of Road Traffic Accidents in the Bangkok Metropolitan Region"
date: "September 19, 2026"
date-modified: "last-modified"
format: html
execute:
  echo: true
  eval: true
  message: false
  warning: false
  freeze: true
editor: visual
---

## 1 Overview

Public safety planning increasingly relies on location-based evidence to identify where and when incidents concentrate, so that enforcement, engineering, and policy responses can be targeted rather than applied uniformly across a region. Road traffic accidents are a point event well suited to this type of analysis: each incident carries a location and a time, and the accumulated pattern of incidents can reveal whether risk is randomly scattered or systematically concentrated in identifiable places and periods.

This exercise applies spatial and spatio-temporal point pattern analysis to road traffic accident data for the [Bangkok Metropolitan Region (BMR)](https://en.wikipedia.org/wiki/Bangkok_Metropolitan_Region), Thailand, covering Bangkok and the five adjacent provinces of Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan, and Samut Sakhon.

## 2 Analytical Questions

1.  How is the intensity of road traffic accidents distributed across space and time within the BMR?
2.  Do accidents exhibit statistically significant spatial and spatio-temporal clustering, beyond what would be expected under complete spatial randomness?
3.  Does accounting for the road network materially change the interpretation of accident concentration, compared to treating accidents as freely distributed in planar space?
4.  What do the resulting patterns imply for public safety planning and resource allocation within the region?

## 3 The Data

Three datasets are used in this exercise:

| Dataset | Description | Format | Source | Date Accessed |
|---------------|---------------|---------------|---------------|---------------|
| Thailand Road Accident (2019 to 2022) | Accident-level records with location, timing, severity, and contributing-factor fields, sourced from the Ministry of Transport's TRAMS dashboard. **Only the 2022 subset is used for this study.** | CSV | [Kaggle](https://www.kaggle.com/datasets/thaweewatboy/thailand-road-accident-2019-2022) | 20 Sep 2026 |
| Thailand Administrative Boundaries, ADM2 | District-level administrative boundaries for Thailand | GeoJSON | [geoBoundaries](https://www.geoboundaries.org) | 20 Sep 2026 |
| Thailand Road Network (OpenStreetMap extract) | Line-based road network for Thailand, classified by highway type under the `fclass` field | Shapefile | [Geofabrik](https://download.geofabrik.de/asia/thailand.html) | 20 Sep 2026 |

## 4 Installing and Loading the R Packages

The following R packages are used across this exercise:

| Package | Purpose | Use Case in This Exercise |
|------|------|------|
| **tidyverse** | Collection of packages for data import, manipulation, and visualisation | Importing, cleaning, and transforming the accident dataset |
| **sf** | Imports, manages, and processes vector-based geospatial data | Handling the administrative boundary and road network data, and constructing spatial point objects from the accident coordinates |
| **spatstat** | Toolkit for spatial point pattern analysis | First-order (KDE, quadrat) and second-order (G, F, K, L functions) analysis of accident intensity and clustering |
| **sparr** | Kernel estimation of spatial and spatio-temporal densities | Spatio-temporal KDE of accident intensity across months in 2022 |
| **stpp** | Analysis of second-order properties of spatio-temporal point processes | Testing for space-time interaction beyond what spatial and temporal clustering alone would predict |
| **spNetwork** | Spatial analysis constrained to a network | Network-constrained kernel density estimation (NKDE) of accidents along the road network |
| **tmap** | Static and interactive thematic mapping | Choropleth maps, KDE surfaces, and accident point maps throughout the report |
| **knitr** | Dynamic report generation | Rendering tables and controlling code chunk output |
| **plotly** | Creates interactive visualisations | Building interactive charts for Exploratory Data Analysis |

```{r}
pacman::p_load(tidyverse, sf, spatstat, sparr, stpp, 
               spNetwork, tmap, knitr, plotly)
```

## 5 Reproducibility

A fixed seed value of 123 is set for reproducibility of any simulation-based procedure used later in this exercise.

```{r}
set.seed(123)
```

## 6 Importing and Preparing the Data

This section imports and prepares each of the three datasets described above. The study area for this exercise is the Bangkok Metropolitan Region (BMR), operationalised at the district (ADM2) level, and the study period is restricted to 2022.

### 6.1 Thailand Road Accident Data

The accident dataset is imported using *read_csv()* of the **readr** package:

```{r}
rdacc <- read_csv("data/thai_road_accident_2019_2022.csv")
```

Null values are checked across every column:

```{r}
colSums(is.na(rdacc))
```

::: callout-note
**Analysis:**

Of the 359 rows with missing coordinates, 350 are attributed to a single agency, the Expressway Authority of Thailand. The other two agencies, Department of Highways and Department of Rural Roads, have almost none missing. The affected rows are concentrated in Bangkok and in the 2021 to 2022 period. These rows cannot be placed spatially, so they are dropped from the point pattern analysis. As a result, accidents attributed to the Expressway Authority, largely Bangkok's controlled-access expressway network, are undercounted relative to the other two agencies.
:::

Exact duplicate rows are checked next:

```{r}
sum(duplicated(rdacc))
```

The set of provinces represented in the data is inspected, since only accidents within the six BMR provinces are relevant to this exercise:

```{r}
sort(unique(rdacc$province_en))
```

The six BMR province names are confirmed to appear exactly as expected in the list above (Bangkok, Nonthaburi, Nakhon Pathom, Pathum Thani, Samut Prakan, Samut Sakhon).

The dataset is filtered to the study area (BMR) and study period (2022), and limited to rows with valid coordinates. It is then converted into a spatial point object.

```{r}
bmr_provinces <- c("Bangkok", "Nonthaburi", "Nakhon Pathom", 
                   "Pathum Thani", "Samut Prakan", "Samut Sakhon")

accidents_bmr <- rdacc %>%
  filter(province_en %in% bmr_provinces,
         year(incident_datetime) == 2022,
         !is.na(latitude), !is.na(longitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = 32647)
```

::: callout-note
- [crs = 4326](https://epsg.io/4326) labels the coordinates as WGS84 decimal degrees
- [crs = 32647](crs%20=%2032647) then reprojects them into UTM Zone 47N (metres). EPSG:32647 specifically covers 96°E to 102°E, the zone the BMR provinces fall within; this reprojection is needed for the distance- and area-based methods used later.
:::

### 6.2 Administrative Boundaries: Bangkok Metropolitan Region (BMR)

Thailand's administrative boundaries run from country (ADM0), through province (ADM1), district (ADM2), to sub-district (ADM3).

This exercise uses both ADM1 and ADM2: ADM1 identifies the six provinces that make up the BMR by name, and ADM2 supplies the actual study area boundary at finer district resolution, needed later for the network-constrained analysis. ADM2's own attribute table does not record which province each district belongs to, so the two levels are combined through a spatial join.

ADM1 is imported first, since it supplies the province names needed to define the BMR.

```{r}
tha_adm1 <- st_read("data/geoBoundaries-THA-ADM1.geojson")
head(tha_adm1)
```

The full list of province names is checked before filtering:

```{r}
sort(unique(tha_adm1$shapeName))
```

Next, ADM2 is imported to supply the actual study area boundary.

```{r}
tha_adm2 <- st_read("data/geoBoundaries-THA-ADM2.geojson")
head(tha_adm2)
```

::: callout-note
`tha_adm2` has 928 features and five attribute fields: shapeName, shapeISO, shapeID, shapeGroup, and shapeType. None of the fields represent a parent province. The BMR district set therefore cannot be filtered by attribute and is derived spatially against ADM1.
:::

Both boundaries are reprojected to EPSG:32647.

```{r}
tha_adm1 <- st_transform(tha_adm1, crs = 32647)
tha_adm2 <- st_transform(tha_adm2, crs = 32647)
```

The six BMR provinces are filtered out of ADM1:

```{r}
bmr_provinces_adm1 <- c("Bangkok", 
                        "Nonthaburi Province",
                        "Nakhon Pathom Province",
                        "Pathum Thani Province",
                        "Samut Prakan Province",
                        "Samut Sakhon Province")

adm1_bmr <- tha_adm1 %>%
  filter(shapeName %in% bmr_provinces_adm1) %>%
  rename(province_name = shapeName)

nrow(adm1_bmr)
```

::: callout-note
`shapeName` is renamed to `province_name` to avoid a duplicate column when joined to `tha_adm2`, which carries its own `shapeName` field for district names.
:::

Each ADM2 district is then assigned to whichever ADM1 province it overlaps most with, and only those districts assigned to one of the six BMR provinces are kept.

```{r}
adm2_bmr <- st_join(tha_adm2, adm1_bmr, 
                    join = st_intersects, 
                    largest = TRUE) %>% 
  filter(!is.na(province_name)) %>%
  select(shapeName, province_name, geometry)
```

```{r}
adm2_bmr %>%
  st_drop_geometry() %>%
  count(province_name)
```

The resulting `adm2_bmr` is the study area boundary used for the remainder of this exercise.

As a visual check, the boundary is plotted and coloured by province. This shows whether the districts form one connected region, or whether any district ended up misplaced or disconnected, something the count above could not reveal.

```{r}
tm_shape(adm2_bmr) +
  tm_polygons(fill = "province_name",
              fill.legend = tm_legend(title = "Province", na.show = FALSE)) +
  tm_layout(main.title = "BMR Study Area by Province",
            main.title.position = "center")
```

### 6.3 Road Network

The road network is imported and inspected:

```{r}
roads <- st_read("data/gis_osm_roads_free_1.shp")
glimpse(roads)
```

```{r}
st_crs(roads)$input
```

The highway classification is inspected via the `fclass` field:

```{r}
sort(unique(roads$fclass))
```

The road network is reprojected to EPSG:32647.

```{r}
roads <- st_transform(roads, crs = 32647)
```

For the purpose of this study, the network is filtered to the intercity highway classes based on the [WikiProject Thailand highway classification convention](https://wiki.openstreetmap.org/wiki/Thailand/Conventions).

```{r}
highway_types <- c("motorway", "motorway_link", 
                   "trunk","trunk_link", 
                   "primary", "primary_link",
                   "secondary", "secondary_link",
                   "tertiary", "tertiary_link", 
                   "unclassified")

roads_filtered <- roads %>%
  filter(fclass %in% highway_types)
```

::: callout-note
The 28 `fclass` values include a `_link` variant for five of the six target classes, representing ramps and interchange connectors rather than a separate road type. These are kept alongside their base class, since dropping them would leave a gap at every interchange, breaking the network needed for the network-constrained analysis later. The rest are excluded as non-vehicular, local, or unclassifiable.
:::

The filtered network is then clipped to the BMR study area:

```{r}
roads_bmr <- st_intersection(roads_filtered, adm2_bmr)
```

Clipping lines against polygons with `st_intersection()` can change geometry types. For example, a line that only touches a polygon boundary at a single point can come out as a point instead of a line. Therefore, the geometry types present in `roads_bmr` are checked.

```{r}
st_geometry_type(roads_bmr) |> unique()
```

The result is a mix of `LINESTRING` and `MULTILINESTRING`.

Since more than one geometry type is present, `roads_bmr` is cast to a single consistent type, `LINESTRING`, which **spNetwork** requires.

```{r}
roads_bmr <- st_cast(roads_bmr, "LINESTRING")
```

The resulting `roads_bmr` is the road network used for the network-constrained kernel density estimation later in this exercise.

## 7 Feature Engineering

`accidents_bmr` currently only has a single timestamp column, `incident_datetime`. This section derives temporal and contextual variables from it, which is needed to examine how accidents vary over time.

**1. Temporal Variables**

Temporal variables are extracted first:

```{r}
accidents_bmr <- accidents_bmr %>%
  mutate(month = month(incident_datetime),
         month_label = month(incident_datetime, label = TRUE, abbr = TRUE),
         day_of_week = wday(incident_datetime, label = TRUE, week_start = 1),
         day_of_year = yday(incident_datetime))
```

**2. Season**

Season is derived based on Thailand's official three-season year - hot, rainy, and cool ([Government Public Relations Department](https://thailand.go.th/event-detail/009_142)):

```{r}
accidents_bmr <- accidents_bmr %>%
  mutate(
    md = format(incident_datetime, "%m-%d"),
    season = case_when(
      md >= "02-15" & md < "05-15" ~ "Hot",
      md >= "05-15" & md < "10-15" ~ "Rainy",
      TRUE ~ "Cool"
    )
  ) %>%
  select(-md)
```

**3. Peak Period**

Peak period is derived based on Bangkok's known morning and evening rush hours. The exact cutoffs used here are a reasonable estimate rather than an official rule.

```{r}
accidents_bmr <- accidents_bmr %>%
  mutate(
    hms = format(incident_datetime, "%H:%M:%S"),
    peak_period = case_when(
      hms >= "07:00:00" & hms <= "09:00:00" ~ "Morning Peak",
      hms >= "16:00:00" & hms <= "19:00:00" ~ "Evening Peak",
      TRUE ~ "Off-Peak"
    )
  ) %>%
  select(-hms)
```

**4. Songkran: "Seven Dangerous Days"**

Songkran's "Seven Dangerous Days" is Thailand's annual road-safety campaign, covering the week surrounding the Songkran New Year holiday, a period nationally recognised for a sharp rise in road accidents. Since this study covers 2022, the relevant campaign dates are 11 to 17 April that year ([The Phuket News](https://www.pressreader.com/thailand/the-phuket-news/20220422/281560884348253)).

```{r}
accidents_bmr <- accidents_bmr %>% 
  mutate(songkran_dangerous_days = 
           incident_datetime >= as.Date("2022-04-11") & 
           incident_datetime < as.Date("2022-04-18"))
```

Finally, unused columns are dropped: `province_th` (same information as `province_en`), `report_datetime` (reporting lag), and `route` (unused free text).

```{r}
accidents_bmr <- accidents_bmr %>%
  select(-province_th, -report_datetime, -route)
```

```{r}
glimpse(accidents_bmr)
```

## 8 Spatial Point Pattern Analysis (SPPA)

This section applies spatial point pattern analysis (SPPA) using **spatstat** to quantify how road traffic accidents are distributed across the BMR, and to test whether the distribution departs from the Complete Spatial Randomness (CSR) hypothesis, which assumes that accidents occur independently and uniformly across the study area.

Three methods are applied:

1.  **First-order analysis:** Examine how accident intensity varies across the region, using kernel density estimation to identify areas of concentration

2.  **Second-order analysis:** Use Ripley's G, F, K, and L functions to test whether accidents are more clustered, more regular, or consistent with CSR at varying distances

3.  **Network-constrained analysis:** Account for accidents being restricted to the road network rather than free to occur anywhere in planar space

### 8.1 Constructing the Point Pattern Objects

Before SPPA can begin, the `sf` objects prepared above are converted into `ppp` (point pattern) and `owin` (observation window) objects, so that they are in the formats **spatstat** works with.

The observation window is built first, from `adm2_bmr`:

```{r}
adm2_owin <- as.owin(adm2_bmr)
adm2_owin
```

The point pattern is then built from `accidents_bmr`, using the above observation window:

```{r}
accidents_ppp <- as.ppp(accidents_bmr, W = adm2_owin)
accidents_ppp
```

Next, the point pattern is checked for duplicated points, since accidents recorded at identical coordinates are treated differently by **spatstat** than by **sf**:

```{r}
any(duplicated(accidents_ppp))
```

::: callout-note
Since no duplicated points are found, `rjitter()` will not be needed.
:::

`accidents_ppp` and `adm2`\_owin are rescaled from metres to kilometres as this keeps the intensity and area readable.

```{r}
accidents_ppp_km <- rescale(accidents_ppp, 1000, "km")
adm2_owin_km <- rescale(adm2_owin, 1000, "km")
```

The rescaled point pattern is plotted against its window, confirming the points fall within the study area:

```{r}
#| fig-width: 12
#| fig-height: 8
plot(unmark(accidents_ppp_km), main = "Road Accidents in the BMR (2022)")
```

### 8.2 First Order Spatial Point Pattern Analysis

First-order analysis begins with kernel density estimation (KDE), which estimates how accident intensity, the expected number of accidents per unit area, varies continuously across the BMR. Unlike a count aggregated by district, KDE produces a smooth surface that does not depend on how the study area happens to be divided into administrative units.

#### 8.2.1 Kernel Density Estimation (KDE)

KDE's most consequential parameter is its bandwidth (`sigma` in **spatstat**), which controls how far each point's influence is smoothed. Too small produces a noisy surface; too large oversmooths and hides genuine concentration. Three automatic selectors, each using a different criterion, are computed and compared rather than fixing a value by inspection.

**Diggle's method** minimises the mean-square error of the estimated intensity and tends to select a small bandwidth sensitive to fine-scale clustering:

```{r}
bw_diggle <- bw.diggle(accidents_ppp_km)
bw_diggle
```

**The Cronie-van Lieshout method** is designed to directly minimise estimation error in the intensity function itself:

```{r}
bw_cvl <- bw.CvL(accidents_ppp_km)
bw_cvl
```

**Likelihood cross-validation** tends toward comparatively larger bandwidths that risk oversmoothing local detail:

```{r}
bw_ppl <- bw.ppl(accidents_ppp_km)
bw_ppl
```

The three values are compared directly:

```{r}
c(diggle = bw_diggle, CvL = bw_cvl, ppl = bw_ppl)
```

The unusually large CvL value, 28.622km against 0.064km and 0.934km for the other two, is further checked.

```{r}
plot(bw_cvl)
```

::: callout-note
**Analysis:**

The curve is flat between roughly 10 km and 28 km before rising sharply. However, the criterion barely distinguishes between bandwidths across that range. At 28.6 km, roughly a quarter of the BMR's extent (90 to 125 km), the result would also collapse local concentration into a single regional trend. For both reasons, `bw.CvL` is not the most appropriate. 
:::

Density surfaces are computed using both remaining bandwidths (`bw.diggle` and `bw.ppl`), for visual comparison before either is treated as final.

```{r}
#| fig-width: 12
#| fig-height: 8

kde_diggle <- density(accidents_ppp_km, sigma = bw_diggle, edge = TRUE, kernel = "gaussian")
kde_ppl <- density(accidents_ppp_km, sigma = bw_ppl, edge = TRUE, kernel = "gaussian")

par(mfrow = c(1, 2))
plot(kde_diggle, main = "Diggle bandwidth")
plot(kde_ppl, main = "ppl bandwidth")
```

::: callout-note
**Analysis:**

Both surfaces trace the same underlying pattern despite using different bandwidths. This agreement is a useful check as it suggests that the pattern reflects real spatial concentration, not an artefact of one bandwidth choice. Diggle's bandwidth shows this pattern at finer resolution and will be used as the primary surface moving forward.
:::

#### 8.2.2 Quadrat Analysis

Quadrat analysis complements KDE with a simpler, non-kernel method. The study area is divided into a grid of equal-sized cells, and the number of accidents in each cell is compared against what Complete Spatial Randomness would predict.

The grid resolution is chosen so that it matches the shape of the study area, and each cell has enough expected accidents for the chi-square test below to be valid; at least 5 per cell is a common rule of thumb. `adm2_owin`'s enclosing rectangle is approximately 177.5 km by 131.1 km, an aspect ratio of about 1.35, which is close to a 12 by 9 grid's ratio of 1.33.

```{r}
qcount <- quadratcount(accidents_ppp_km, nx = 12, ny = 9)

plot(unmark(accidents_ppp_km), pch = 20, cex = 0.4, main = "Quadrat Counts of Road Accidents")
plot(qcount, add = TRUE, col = "red", cex = 0.8)
```

The variance to mean ratio (VMR) provides a quick diagnostic. A value close to 1 is consistent with CSR, above 1 suggests clustering, and below 1 suggests regularity. 

```{r}
qcount_df <- as.data.frame(qcount)
vmr <- var(qcount_df$Freq) / mean(qcount_df$Freq)
vmr
```

A chi-square test of CSR is performed, using the same grid:

```{r}
quadrat.test(accidents_ppp_km, nx = 12, ny = 9)
```

::: callout-note
**Analysis:**

VMR and the chi-square test answer different questions: VMR measures how large the departure from CSR is; the chi-square test measures whether that departure could be due to chance. With a sample this large, the chi-square test is almost always significant and the test alone would not show the effect is substantial. VMR fills this gap: at 176.6, where the value is significantly higher than 1 expected under CSR, it shows the clustering is genuinely large. The chi-square test confirms it is not due to chance. This agrees with the KDE surface, both methods point to the same conclusion.
:::

#### 8.2.3 Temporal First-Order Analysis

KDE and quadrat analysis above addressed how accident intensity varies across space. This section focuses on the temporal portion, examining whether accident intensity varies across the year. The point pattern is split by month, and a separate density surface is computed for each.

```{r}
accidents_ppp_month <- accidents_ppp_km
marks(accidents_ppp_month) <- marks(accidents_ppp_km)$month

st_kde <- spattemp.density(accidents_ppp_month)
summary(st_kde)
```

Twelve monthly slices are plotted on a shared colour scale for direct comparison across the year:

```{r}
#| fig-width: 10
#| fig-height: 10

par(mfrow = c(4, 3), mar = c(2, 2, 2, 2))
for (i in 1:12) {
  plot(st_kde, i,
       override.par = FALSE,
       fix.range = TRUE,
       main = paste("Month", i))
}
```

::: callout-note
**Analysis:**

**Month 1** shows a moderately widespread hotspot concentrated in the central BMR, with some diffusion into surrounding areas.

**Months 2 to 7** show generally lower intensity, with hotspots more contained.

**Months 8 to 11** show accident density concentrated around Bangkok increasing in both intensity and spatial extent, with the branching pattern along major roads becoming more pronounced.

**Month 12** shows the highest density of the year, concentrated in central BMR, likely due to increased travel during the year-end and holiday period. This also coincides with the start of Thailand's New Year "Seven Dangerous Days" period, 29 December 2022 to 4 January 2023, of which 29 to 31 December falls within the study period. 317 deaths were recorded nationwide [GT Rider, citing Thailand's Road Safety Centre, 2023](https://www.thegtrider.com/thailand-2022-2023-new-year-road-traffic-report-deaths.htm).
:::

#### 8.2.4 KDE Comparison By Province

The region-wide KDE surface shows where accidents concentrate across the whole BMR, but not how that pattern differs between provinces. This section computes a separate density surface for each of the six BMR provinces, using each province's own boundary as the window.

```{r}
kde_by_province <- list()

for (i in seq_along(bmr_provinces)) {
  province_owin <- adm2_bmr %>%
    filter(province_name == bmr_provinces_adm1[i]) %>%
    as.owin() %>%
    rescale(1000, "km")

  province_ppp <- accidents_ppp_km[province_owin]
  kde_by_province[[bmr_provinces[i]]] <- density(province_ppp, sigma = bw_diggle, edge = TRUE, kernel = "gaussian")
}
```

The six surfaces are plotted on a shared colour scale for direct comparison:

```{r}
#| fig-width: 12
#| fig-height: 12
province_range <- range(sapply(kde_by_province, function(im) range(im$v, na.rm = TRUE)))

par(mfrow = c(3, 2), mar = c(2, 2, 2, 4))
for (nm in names(kde_by_province)) {
  plot(kde_by_province[[nm]], main = nm, axes = TRUE, box = TRUE)
}
```

::: callout-note
**Analysis:**

Across all six provinces, accidents cluster tightly along specific road corridors, appearing on each map as thin, bright lines.

**Bangkok** and **Samut Prakan** show the highest peak intensities among the six provinces, with colour scales reaching approximately 170 and 250 respectively. In both cases, this intensity is concentrated along short, specific highway segments rather than distributed across the province. **Nakhon Pathom** shows the lowest peak intensity, with its colour scale reaching only about 10, and its hotspots appear comparatively scattered rather than concentrated on one corridor. **Nonthaburi** and **Pathum Thani** fall in between, with colour scales reaching approximately 44 and 50 respectively. For both provinces, hotspots concentrate toward the southern portion of the province, the area bordering Bangkok.
:::

## Reference
